# Linear Correlation: "Cancer" 


```{r, include=FALSE}
# set global chunk options...  
#  this changes the defaults so you don't have to repeat yourself
knitr::opts_chunk$set(comment     = NA,
                      cache       = TRUE,
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      out.width   = "75%")
```



```{r, comment=FALSE, message=FALSE}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(haven)        # Read in SPSS datasets
library(psych)        # Lots of nice tid-bits
library(GGally)       # Extension to 'ggplot2' (ggpairs)
library(corrplot)     # Correlation Matrix plots
library(furniture)    # tables
library(apaSupp)      # APA 7th ed tables
```

## PURPOSE

### Research Question

> Is there a relationship between the demographics and oral condition over the 6 weeks?


### Data Description

The `Cancer` data set contains 4 repeated measure of oral condition for 25 cancer patients to see if an aloe treatment helps with oral condition (ulcers).



```{r}
cancer_raw <- haven::read_spss("https://raw.githubusercontent.com/CEHS-research/eBook_ANOVA/master/data/Cancer.sav")

tibble::glimpse(cancer_raw)
```

```{r}
cancer_clean <- cancer_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(trt = factor(trt,
                             labels = c("Placebo", 
                                        "Aloe Juice"))) %>% 
  dplyr::mutate(stage = factor(stage))

tibble::glimpse(cancer_clean)
```

```{r df_cancer}
psych::headTail(cancer_clean) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Sample of Cancer Dataset")
```




## EXPLORATORY DATA ANALYSIS

### Scatterplot

```{block type='rmdimportant', echo=TRUE}
Always plot your data first!
```

```{r}
cancer_clean %>% 
  ggplot(aes(x = age,
             y = weighin)) +
  geom_count() +
  geom_smooth(method = "lm") +
  theme_bw()
```




## PEARSON'S CORRELATION

### Using the Default Settings


```{block type='rmdlightbulb', echo=TRUE}
The `cor.test()` function needs at least TWO arguments:

* **formula** - The formula specifies the two variabels between which you would like to calcuate the correlation.  Note at the two variable names come AFTER the tilda symbol and are separated with a plus sign: `~ continuous_var1 + continuous_var2`

* **data** - Since the datset is not the first argument in the function, you must use the period to signify that the datset is being piped from above `data = .` 
```

```{r}
cancer_clean %>% 
  cor.test(~ age + weighin,  # formula: order doesn't matter
           data = .)         # data piped from above
```




### Additional Arguments


```{block type='rmdlightbulb', echo=TRUE}
* **alternative**  - The `cor.test()` function defaults to the `alternative = "two.sided"`.  If you would like a one-sided alternative, you must choose which side you would like to test:  `alternative = "greater"` to test for POSITIVE correlation or `alternative = "less"` to test for NEGATIVE correlation.

* **method** - The default is to calculate the Pearson correlation coefficient (`method = "pearson"`), but you may also specify the Kendall's tau (`method = "kendall"`)or Spearman's rho (`method = "spearman"`), which are both non-parametric methods.  

* **conf.int** - It also defaults to testing for the two-sided alternative and computing a 95\% confidence interval (`conf.level = 0.95`), but this may be changed.  
```

Since the following code only specifies the**defaults**, it Will give the same results as if you did not type out the last three lines (see above).

```{r}
cancer_clean %>% 
  cor.test(~ age + weighin,
           data = .,
           alternative = "two.sided",  # or "greater" (positive r) or "less" (negative r)
           method = "pearson",          # or "kendall" (tau) or "spearman" (rho)
           conf.level = .95)            # or .90 or .99 (ect)
```

**Non-Significant Correlation**



> **APA Results:** There was no evidence of an association in overall oral condition from baseline to two week follow-up, $r(25) = -0.288 $p < .163$.



### Statistical Significance

```{r}
cancer_clean %>% 
  ggplot(aes(x = totalcin,
             y = totalcw4)) +
  geom_count() +
  geom_smooth(method = "lm") +
  theme_bw()
```

```{r}
cancer_clean %>% 
  cor.test(~ totalcin + totalcw4,
           data = .)
```



**Statistically Significant Correlation**

> **APA Results:** Overall oral condition was positively correlated ($r = .763$) between weeks two and four, $t(21) = 5.409$, $p < .001$.



```{r}
cancer_clean %>% 
  ggplot(aes(x = totalcw4,
             y = totalcw6)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```


```{r}
cancer_clean %>% 
  cor.test(~ totalcw4 + totalcw6,
           data = .)
```




## CORRELATION TABLES


```{block type='rmdlightbulb', echo=TRUE}
The may use the `tableC()` function from the `furniture` package to calculate all pair-wise correlations between more than two variables and arrange them all in a table.  The table is formatted with the variabels listed on the rows and numbered to show the same variabels across the columns. 

* The cells ON the diagonal are all equal to *exactly one*, since each variable is perfectly correlated with itself.

* The cells ABOVE the diagonal are *blank* as them would just be a mirror image of the values below the diagonal. 

* The cells BELOW the diagonal each contain the *Pearson's correlation coefficients* for each pair of variables, $r$, with the $p-value$ showing the significance vs. the null hypothesis for no association ($r = 0$) to the right.
```

```{r}
cancer_clean %>% 
  furniture::tableC(age, weighin, totalcin)
```


### Missing Values - Default


```{block type='rmdlightbulb', echo=TRUE}
**Default Behavior `na.rm = FALSE` (default)**

If you don't say otherwise, the correlation $r$ with not be calculated (`NA`) between any pair of variables for which there is at least one subject with a missing value on at least one of the vairables.  This is a nice alert to make you aware of missing values.
```

```{r}
cancer_clean %>% 
  furniture::tableC(totalcin, totalcw2, totalcw4, totalcw6)
```

### Missing Values - Listwise Deletion


```{block type='rmdlightbulb', echo=TRUE}
**Listwise Deletion `na.rm = TRUE`**

Most of the time you will want to compute the correlation $r$ is the precense of missing values.  To do so, you want to remove or exclude subjects with missing data from ALL correlation computation in the table.  This is called **'list-wise deletion'**.  It ensures that all cells in the table refer to the exact same sub-sample (n = subjects with complete data for all variables in the table), and thus the same degrees of freedom (since $df = n - 2$).  This is done be changing the default to `na.rm = TRUE`.
```

```{r}
cancer_clean %>% 
  furniture::tableC(totalcin, totalcw2, totalcw4, totalcw6,
                    na.rm = TRUE)
```



```{r}
cancer_clean %>% 
  furniture::tableC(totalcin, totalcw2, totalcw4, totalcw6,
                    na.rm = TRUE,
                    rounding = 2)
```


### APA Style

```{r}
cancer_clean %>% 
  dplyr::select("Baseline" = totalcin, 
                "Week 2" = totalcw2, 
                "Week 4" = totalcw4, 
                "Week 6" = totalcw6) %>% 
  apaSupp::tab_cor(caption = "Correlations Between Repeated Measures of Total Oral Condition")
```



## PAIRS PLOTS

* [Helpful Website](http://jamesmarquezportfolio.com/correlation_matrices_in_r.html)



### Using Base R

```{r, fig.asp=1}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  pairs()
```





### Using the `psych` package

```{r, fig.asp=1}
cancer_clean %>% 
  dplyr::select(age, weighin,
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  psych::pairs.panels()
```




### Using the `ggplot2` and `GGally` packages


```{r, fig.asp = 1}
cancer_clean %>% 
  dplyr::select(age, weighin,
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  data.frame %>% 
  ggscatmat() +
  theme_bw()
```


```{r, fig.asp = 1}
cancer_clean %>% 
  data.frame %>% 
  ggscatmat(columns = c("age", "weighin",
                        "totalcin", "totalcw2", "totalcw4", "totalcw6"),
            color = "trt") +
  theme_bw()
```






## CORROLAGRAMS

### Simple Correlation Matrix (Base R)

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") 
```

### Using the Default Settings

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") %>% 
  corrplot::corrplot()
```


### Changing the shape

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") %>% 
  corrplot::corrplot(method = "square")
```

### Only Displaying Half

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") %>% 
  corrplot::corrplot(method = "ellipse",
                     type = "lower")
```

### Mixing it up

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") %>% 
  corrplot::corrplot.mixed()
```

### Getting Fancy

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, 
                totalcin, totalcw2, totalcw4, totalcw6) %>% 
  cor(method = "pearson",
      use = "complete.obs") %>% 
  corrplot::corrplot.mixed(upper = "number",
                           lower = "ellipse")
```







