
# Simple Regression: "Minigolf"


Darlington & Hayes, Chapter 2



```{r}
# library(remotes)
# remotes::install_github("sarbearschwartz/apaSupp")

library(tidyverse)
library(flextable)
library(apaSupp)
library(olsrr)
library(modelsummary)
library(parameters)
library(effectsize)
library(interactions)
library(performance)
library(ggResidpanel)
library(jtools)
library(olsrr)
library(ggpubr)
```


## PURPOSE

### Research Question

> Examine the relationship between the number of points won when playing mini golf and the number of times a player has played mini golf before before.


### Data Description

Manually enter the data set provided on page 18 in Table 2.1

```{r}
df_golf <- data.frame(ID = 1:23,
                      X = c(0, 0, 
                            1, 1, 1, 
                            2, 2, 2, 2, 
                            3, 3, 3, 3, 3, 
                            4, 4, 4, 4,
                            5, 5, 5, 
                            6, 6),
                      Y = c(2, 3, 
                            2, 3, 4,
                            2, 3, 4, 5,
                            2, 3, 4, 5, 6,
                               3, 4, 5, 6,
                                  4, 5, 6,
                                     5, 6)) %>% 
  dplyr::mutate(ID = as.character(ID)) %>% 
  dplyr::mutate(across(c(X, Y), as.integer))  
```


```{r}
tibble::glimpse(df_golf)
```



\clearpage

## VISUALIZATION

### Full Table


This is Table 2.1 on page 18.

```{r}
tab_df_golf <- df_golf %>% 
  dplyr::select(ID,
                "X\n'Previous Plays'" = X,
                "Y\n'Points Won'" = Y) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H Table 2.1: Golfing Scores and Prior Plays") %>% 
  flextable::colformat_double(digits = 0)
```


This is how you save a table to Word.

```{r}
flextable::save_as_docx(tab_df_golf,
                        path = "tables/tab_df_golf.docx")
```



\clearpage


```{r}
tab_df_golf
```


\clearpage

### Scatterplot

This is Figure 2.1 on page 19

```{r, collapse = TRUE}
fig_blank <- df_golf %>% 
  ggplot(aes(x = X,
             y = Y)) +
  geom_vline(xintercept = 0,
             linewidth = 1) +
  geom_hline(yintercept = 0,
             linewidth = 1) +
  scale_x_continuous(breaks = 0:6, limits = c(-.5, 6.5)) +
  scale_y_continuous(breaks = 0:6, limits = c(-.5, 6.5)) +
  labs(x = "X = Number of Previous Plays",
       y = "Y = Points Won") +
  theme_bw() +
  theme(panel.grid.minor = element_line(linewidth = .5, 
                                        linetype = "longdash"),
        panel.grid.major = element_line(linewidth = 1))
```


```{r, collapse = TRUE}
fig_golf_scatter <- fig_blank +
  geom_point(shape = 15,
             size = 4) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) 
```



```{r, fig.cap="D&H Figure 2.1 (page 19) A Simple Scatter Plot"}
fig_golf_scatter 
```

\clearpage

This is how you save a plot, in three different formats.

```{r}
ggsave(plot = fig_golf_scatter,
       file = "figures/fig_golf_scatter.png",
       width = 6,
       height = 4,
       units = "in")
```

```{r}
ggsave(plot = fig_golf_scatter,
       file = "figures/fig_golf_scatter.jpg",
       width = 6,
       height = 4,
       units = "in")
```


```{r}
ggsave(plot = fig_golf_scatter,
       file = "figures/fig_golf_scatter.tif",
       width = 6,
       height = 4,
       units = "in")
```

\clearpage

### Conditional Means


```{r}
df_golf_means <- df_golf %>% 
  dplyr::group_by(X) %>% 
  dplyr::summarise(N = n(),
                   Y_bar = mean(Y)) %>% 
  dplyr::ungroup()
```

 
```{r, fig.cap="Conditional Mean of Y When X = 0"}
fig_blank +
  geom_rect(xmin = 0 - .25, xmax = 0 + 0.25,
            ymin = 2 - .25, ymax = 3 + 0.25,
            color = "red",
            alpha = .5,
            fill = "yellow") +
  geom_point(shape = 15,
             size = 4) +
  geom_point(data = df_golf_means %>% 
               dplyr::filter(X <= 0),
             aes(x = X,
                 y = Y_bar),
             color = "red",
             shape = 13,
             size = 10) 
```

\clearpage


```{r, fig.cap="Conditional Mean of Y When X = 1"}
fig_blank +
  geom_rect(xmin = 1 - .25, xmax = 1 + 0.25,
            ymin = 2 - .25, ymax = 4 + 0.25,
            color = "red",
            alpha = .5,
            fill = "yellow") +
  geom_point(shape = 15,
             size = 4) +
  geom_point(data = df_golf_means %>% 
               dplyr::filter(X <= 1),
             aes(x = X,
                 y = Y_bar),
             color = "red",
             shape = 13,
             size = 10)
```


\clearpage



```{r, fig.cap="Conditional Mean of Y When X = 2"}
fig_blank +
  geom_rect(xmin = 2 - .25, xmax = 2 + 0.25,
            ymin = 2 - .25, ymax = 5 + 0.25,
            color = "red",
            alpha = .5,
            fill = "yellow") +
  geom_point(shape = 15,
             size = 4) +
  geom_point(data = df_golf_means %>% 
               dplyr::filter(X <= 2),
             aes(x = X,
                 y = Y_bar),
             color = "red",
             shape = 13,
             size = 10)
```

\clearpage



```{r}
df_golf_means %>% 
  dplyr::select("X\nPrevious\nPlays" = N,
                "N\nNumber of\nObservations" = N,
                "Mean(Y)\nConditional Mean of\nPoints Won" = Y_bar) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Golfing Scores and Prior Plays")  %>% 
  flextable::align(part = "body", align = "center") %>% 
  flextable::align(part = "head", align = "center")
```


\clearpage



```{r, fig.cap="Textbook's Figure 2.2 (page 20) A Line Through Conditional Means"}
fig_golf_scatter +
  geom_point(data = df_golf_means,
             aes(x = X,
                 y = Y_bar),
             color = "red",
             shape = 13,
             size = 10)
```


\clearpage



```{r, fig.cap="Y-Intercept and Slope of the Line"}
fig_blank +
  geom_point(shape = 15,
             size = 4,
             alpha = .3) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  geom_point(x = 0,
             y = 2.5,
             color = "red",
             shape = 13,
             size = 10) +
  geom_segment(x = -.75, xend = -.1,
               y = 2.5, yend = 2.5,
               arrow = arrow(type = "closed"),
               color ="red") +
  geom_segment(x = 1, xend = 1,
               y = 3, yend = 4,
               arrow = arrow(length = unit(.3, "cm"),
                             type = "closed"),
               linewidth = 1,
               color = "darkgreen") +
  geom_segment(x = 1, xend = 3,
               y = 4, yend = 4,
               arrow = arrow(length = unit(.3, "cm"),
                             type = "closed"),
               linewidth = 1,
               color = "darkgreen") +
  annotate(x = 0.75, y = 3.5, 
           geom = "text", 
           label = "Rise 1",
           color = "darkgreen")+
  annotate(x = 2, y = 4.5, 
           geom = "text", 
           label = "Run 2",
           color = "darkgreen") +
  ggpubr::stat_regline_equation(label.x = 4.5,
                                label.y = 1,
                                size = 6)
```

### Y-intercept

Where the line crosses the *vertical* or y-axis

$$
b_0 = 2.5 
$$

### Slope

The incline or decline of the line

$$
b_1 = \frac{rise}{run} = \frac{1}{2} = 0.5
$$


\clearpage

### Format

Standard slope-intercept form


$$
Y = mX + b \\
\text{or} \\
Y = b + mX
\tag{Slope-Intercept Form}
$$

In statistics:

$$
Y = b_0 + b_1X
\tag{D&H 2.10}
$$

So for this example, $\hat{Y}$ is said "Y hat".

$$
\hat{Y} = 2.5 + 0.5X \\
\text{or} \\
\widehat{\text{points}} = 2.58 + 0.5(\text{plays}) 
$$


\clearpage


## HAND CALCULATIONS


ORDINARY LEAST SQUARES (OLS)

### Estimates

The equation may be used to estimate predicted values ($\hat{Y}$) for each participant ($i$).  

$$
\tag{OLS EQ}
\widehat{Y_i} = 2.5 + 0.5X_i
$$


The first participant (`id = 1`) had no previous plays ($X = 0$) and won two points ($Y=2$)...

```{r}
df_golf %>% 
  dplyr::filter(ID == "1")
```

...so we plug in the value of `0` for the variable $x$ in the OLS Equation...

$$
\hat{Y} = 2.5 + 0.5 (0) \\
= 2.5 + 0 \\
= 2.5 \\
$$

...which gives a predicted value of two and a half points won ($\hat{Y} = 2.5$).


\clearpae

### Residuals

The words `error` and `residual` mean the same thing ($e$).  

$$
\tag{residuals}
residual  = \text{observed} - \text{predicted} \\
\text{or} \\
e_i= Y_i - \widehat{Y_i}
$$
For the first participant (`ID = 1`), this would be...

$$
e_1 = (2 - 2.5) 
= - 0.5
$$

This is because this participant won two points, which is a half a point LESS THAN the OLS equation which predicted two and a half points.


\clearpage


We can use this process to find the predicted values ($\hat{Y}$) and residuals ($e$)  for all the participants in this sample (`N = 23`). 

```{r}
mean(df_golf$X)

mean(df_golf$Y)
```



```{r}
df_golf_est <- df_golf %>% 
  dplyr::mutate(X2 = X^2) %>% 
  dplyr::mutate(Y2 = Y^2) %>% 
  dplyr::mutate(devX = X - 3) %>% 
  dplyr::mutate(devY = Y - 4) %>% 
  dplyr::mutate(devX_devY = devX*devY) %>% 
  dplyr::mutate(devX2 = devX^2) %>% 
  dplyr::mutate(devY2 = devY^2) %>%
  dplyr::mutate(estY = 2.5 + 0.5*X) %>%  # predicted
  dplyr::mutate(e = Y - estY) %>%        # deviation or residual
  dplyr::mutate(e2 = e^2)
```


```{r}
df_golf_est_sums <- df_golf_est %>% 
  dplyr::summarise(across(where(is.numeric),~sum(.x))) %>% 
  dplyr::mutate(ID = "Sum") %>% 
  dplyr::select(ID, everything())
```

```{r}
df_golf_est_means <- df_golf_est %>% 
  dplyr::summarise(across(where(is.numeric),~mean(.x))) %>% 
  dplyr::mutate(ID = "Mean") %>% 
  dplyr::select(ID, everything())
```


```{r}
tab_golf_est <- df_golf_est %>% 
  dplyr::bind_rows(df_golf_est_sums) %>% 
  dplyr::bind_rows(df_golf_est_means) %>% 
  dplyr::select(ID, X, Y,estY, e, e2) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H Tabel 2.2: Estimates and Residuals") %>% 
  flextable::hline(i = 23) 
```

\clearpage

```{r}
tab_golf_est
```


\clearpage

### Errors of Estimate

"Sum of the Squared Residuals" or "Sum of the Squared Errors" ($SS_{residual}$)

$$
SS_{residual} = \sum^{N}_{i = 1}{(Y_i - \hat{Y_i})^2 = \sum^{N}_{i = 1}{e_i^2}}
\tag{D&H 2.1}
$$

For this golf example, $SS_{residual}$ = 25.00.


\clearpage

### Deviation Scores

**Deviation Scores** measures how far an observed value is from the MEAN of all observed values for that variable.


$$
\tag{deviation}
\text{deviation}  = \text{observed} - \text{mean} \\ 
$$

Deviations may be calculated for each variable, separately.

> Note: In our textbook lower cases letters here represent the deviation scores of the larger letter counterparts.


$$
x_i= X_i - \bar{X_i}\\
y_i= Y_i - \bar{Y_i} 
$$



\clearpage


### Cross-Products & Squares

**Cross-Product** is another term for multiply, specifically when talking about the the **deviance scores**.


```{r}
df_golf_est %>% 
  dplyr::bind_rows(df_golf_est_sums) %>% 
  dplyr::bind_rows(df_golf_est_means) %>% 
  dplyr::select(ID, X, Y, 
                # "Squared\nX" = X2, 
                # "Squared\nY" = Y2,
                "Deviation\nX" = devX, 
                "Deviation\nY" = devY, 
                "Squared\nDeviation\nof X" = devX2,
                "Squared\nDeviation\nof Y" = devY2,
                "Deviation\nCross\nProduct" = devX_devY) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H Table 2.3: Regression Computations") %>% 
  flextable::hline(i = 23)
```


\clearpage



### Covariance

Usually not interpreted, but useful in computing things in stats, such as the regression coefficients and correlations.

> Covariance is a measure of the **relationship between two random variables** and to what extent, they **change together**. Or we can say, in other words, it defines the changes between the two variables, *such that change in one variable is equal to change in another variable*. This is the property of a function of maintaining its form when the variables are linearly transformed. Covariance is measured in units, which are calculated by multiplying the units of the two variables.


Note: $N$ is the size of the sample, or the number of observations


The formula for **COVARIANCE** in a full known population (size = `N`) is:

$$
Cov(XY) = \frac{\sum_{i = 1}^{N}{(X - \bar{X})(Y - \bar{Y})}}{N}
\tag{D&H 2.2}
$$

So for our golf example:

```{r}
34/23
```







\clearpage

### Important Note

> D&H Textbook: starting in the middle of page 27!

```{block type='rmdlightbulb', echo=TRUE}
There are two *slightly different variations* to the covariance and variance formulas.  The one for covariance above (EQ 2.2) showns dividing by `N`and is specific to a POPULATION ($\sigma_{XY}$).  

When the data used is a SAMPLE, and you are ESTIAMTING the POPULATION PARAMETER, you divide by `n - 1` rather than `N`.  
```

```{r}
34/23         # divide by N
34/(23 - 1)   # divide by n - 1
```

This is important because in `R`, the functions ASSUME you have a SAMPLE, not a population.

```{r}
cov(df_golf$X, df_golf$Y)  # does the sample version
```



\clearpage


### Covar to Var

This is another version of Equation 2.2 that is a bit more 'complex', but can be helpful.

$$
Cov(XY) = \frac{N\sum_{i = 1}^{N}{X_iY_i} - 
(\sum_{i = 1}^{N}{X_i})(\sum_{i = 1}^{N}{Y_i})}{N^2}
$$

Variance is the same as the Covariance of a variable with itself.  

$$
Cov(XX) = \frac{N\sum_{i = 1}^{N}{X_iX_i} - 
(\sum_{i = 1}^{N}{X_i})(\sum_{i = 1}^{N}{X_i})}{N^2}
$$

Now we can simplify.

$$
Var(X) = \frac{N\sum_{i = 1}^{N}{X_i^2} - 
(\sum_{i = 1}^{N}{X_i})^2}{N^2} 
$$


And some more...

$$
Var(X) = \frac{\sum_{i = 1}^{N}{(X_i - \bar{X_i})^2}}{N}
\tag{D&H 2.3}
$$









\clearpage

### Variance

The formula for **VARIANCE** in a full known population (size = `N`) is:

$$
Var(X) = \frac{\sum_{i = 1}^{N}{(X_i - \bar{X_i})^2}}{N}
\tag{D&H 2.3}  
$$


So for our golf example, variable X:


```{r}
68/23
68/22
```

```{r}
var(df_golf$X)
```

So for our golf example, variable Y:

```{r}
42/23
42/22
```


```{r}
var(df_golf$Y)
```





\clearpage

### Standard Deviation

```{block type='rmdlightbulb', echo=TRUE}
Instead of interpreting variance, we usually refer to **STANDARD DEVIATION**.
```


$$
SD_X = \sqrt{Var(X)}
$$

So for our golf example, variable X:

```{r}
sqrt(68/23)
sqrt(68/22)
```

```{r}
sd(df_golf$X)
```

So for our golf example, variable Y:

```{r}
sqrt(42/23)
sqrt(42/22)
```

```{r}
sd(df_golf$Y)
```


\clearpage

### Correlation

Pearson Product-Moment Correlation coefficient:


$$
r_{XY} = \frac{Cov(XY)}{SD_X \times SD_Y}
\tag{D&H 2.4}
$$

So for our golf example:

```{r}
(34/23)/(sqrt(68/23)*sqrt(42/23))
```


```{r}
cor(df_golf$X, df_golf$Y)
```


\clearpage

### Coefficient or Slope

Covariance can be used to find the SLOPE:

$$
b_1 = \frac{Cov(XY)}{Var(X)}
\tag{D&H 2.5}
$$

For our golf example:

```{r}
cov(df_golf$X, df_golf$Y)/var(df_golf$X)
```



but I prefer this formula that uses summary statistics.


$$
b_1 = r\frac{SD_Y}{SD_X}
\tag{D&H 2.6}
$$

For our golf example:

```{r}
cor(df_golf$X, df_golf$Y)*(sd(df_golf$Y)/sd(df_golf$X))
```


\clearpage

### Constant or Y-intercept

$$
b_0 = \bar{Y} - b_1 \bar{X}
\tag{D&H 2.8}
$$

```{r}
mean(df_golf$Y) - 0.5*mean(df_golf$X)
```


### Regression Equation


$$
Y = b_0 + b_1X
\tag{D&H 2.10}
$$

So for this example, $\hat{Y}$ is said "Y hat".

$$
\hat{Y} = 2.5 + 0.5X \\
\text{or} \\
\widehat{\text{points}} = 2.58 + 0.5(\text{plays}) 
$$

\clearpage


## USING SOFTWARE

### Linear Model

* The dependent variable (DV) is points won ($Y$)
* The independent variable (IV) is number of time previously played ($X$)


```{r}
fit_lm_golf <- lm(Y ~ X,
                  data = df_golf)
```

```{r}
summary(fit_lm_golf)
```



```{r}
olsrr::ols_regress(fit_lm_golf)
```






### Coefficients - Raw

Slope and intercept

```{r}
broom::tidy(fit_lm_golf) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Linear Regression Coefficients")
```



```{r}
coef(fit_lm_golf)
```




```{r}
jtools::summ(fit_lm_golf)
```

### Coefficients - Standardized

```{r}
lm(scale(Y) ~ scale(X),
   data = df_golf) %>% 
  summary()
```






```{r}
cor(df_golf$X, df_golf$Y)
```

```{r}
cor(df_golf$Y, df_golf$X)
```


```{r}
broom::glance(fit_lm_golf) 
```

```{r}
broom::glance(fit_lm_golf)$r.squared %>% sqrt()
```



\clearpage

### Residuals

```{r}
round(fit_lm_golf$residuals, 2)
```



```{r}
broom::augment(fit_lm_golf) 
```


\clearpage

### Errors of Estimate

"Sum of the Squared Residuals" or "Sum of the Squared Errors" ($SS_{residuals}$)

```{r}
sum(fit_lm_golf$residuals^2) 
```


```{r}
anova(fit_lm_golf)
```




\clearpage

### Visualize

```{r, fig.cap="Regress Points Won on Number of Previous Plays"}
df_golf %>% 
  ggplot(aes(x = X,
             y = Y)) +
  theme_bw() +
  geom_point(size = 4,
             alpha = .4) +
  geom_smooth(method = "lm",
              formula = y ~ x)  +
  annotate(x = 0.5,
           y = 5.5,
           size = 6,
           geom = "text",
           label = "r = .636") +
  ggpubr::stat_regline_equation(label.x = 0,
                                label.y = 6,
                                size = 6) +
  labs(x = "Number of Previous Plays",
       y = "Points Won")
```


\clearpage

## RESIDUALS

### Properties

1. Mean of residuals = zero

```{r}
summary(fit_lm_golf$residuals)
```


2. Zero correlation between residuals the X


```{r}
cor(fit_lm_golf$residuals, df_golf$X)
```

3. Variance of residuals = Proportion of Variance not Explained

$$
\frac{Var(\text{residuals})}{Var(Y)} = 1 - r^2
\tag{D&H 2.12}
$$


```{r}
var(fit_lm_golf$residuals)/var(df_golf$Y)
```

```{r}
1 - (cor(df_golf$X, df_golf$Y))^2
```


\clearpage


### Residual Analysis



```{r}
ggResidpanel::resid_xpanel(fit_lm_golf)
```

```{r}
ggResidpanel::resid_panel(fit_lm_golf)
```

```{r}
performance::check_residuals(fit_lm_golf)
```


```{r}
olsrr::ols_plot_diagnostics(fit_lm_golf)
```


