
# Multiple Regression: "Weight Loss"


Darlington & Hayes, Chapter 3's first example

> Darlington, Richard B., and Andrew F. Hayes. Regression Analysis and Linear Models : Concepts, Applications, and Implementation, Guilford Publications, 2016. ProQuest Ebook Central, http://ebookcentral.proquest.com/lib/usu/detail.action?docID=4652287.
Created from usu on 2025-01-29 01:13:21.

```{block type='rmdlink', echo=TRUE}
Outside resource
* https://www.harshaash.com/R/part-and-partial-corr/
* https://dstanley4.github.io/psyc4780bookdown/multiple-regression.html

Reference:

* Knapp, T. R. (1996). [Semi-Partial Correlations: I Don't Need Them; You Can Have Them.](https://scholarworks.bgsu.edu/cgi/viewcontent.cgi?article=1631&context=mwer) Mid-Western Educational Researcher, 9(4), 4.
```


```{r}
# library(remotes)
# remotes::install_github("sarbearschwartz/apaSupp")

library(tidyverse)
library(flextable)
library(apaSupp)
library(olsrr)
library(modelsummary)
library(parameters)
library(effectsize)
library(interactions)
library(performance)
library(ggResidpanel)
```




## PURPOSE

### Research Question

> How is weight loss associationed with exercise frequency and food intake? 


### Data Description

Suppose you conducted a study examining the relationship between food consumption and weight loss among people enrolled (`n` = 10) in a month-long healthy living class.

* `exer` average weekly hours of exercise
* `diet` average daily food consumption (in 100s of calories about the recommended minimum of 1,000 calories required to maintain good health)
* `loss` average weight loss in hundreds of grams per week



```{block type='rmdlightbulb', echo=TRUE}
Manually enter the data set provided on page 44 in Table 3.1
```


```{r}
df_loss <- tibble::tribble(~id, ~exer, ~diet, ~meta, ~loss,
                            1, 0, 2, 15,  6, 
                            2, 0, 4, 14,  2,
                            3, 0, 6, 19,  4,
                            4, 2, 2, 15,  8,
                            5, 2, 4, 21,  9,
                            6, 2, 6, 23,  8,
                            7, 2, 8, 21,  5,
                            8, 4, 4, 22, 11,
                            9, 4, 6, 24, 13,
                           10, 4, 8, 26,  9)
```



```{r}
df_loss %>% 
  dplyr::select("ID" = id,
                "Exercise\nFrequency" = exer,
                "Food\nIntake" = diet,
                "Weight\nLoss" = loss) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H Table 3.1: Exercise, Food Intake, and Weight Loss") %>% 
  flextable::colformat_double(digits = 0) %>% 
  flextable::footnote(part = "header", 
                      i = 1, j = 2:4, 
                      value = as_paragraph(c("Weekly average of hours/day",
                                             "Daily average of 100s of calories above recommendation",
                                             "Weekly average of 100s of grams/week")))
```



## EXPLORATORY DATA ANALYSIS

### Summary Statistics

```{r}
df_loss %>% 
  dplyr::select("Exercise Frequency" = exer,
                "Food Intake" = diet,
                "Weight Loss" = loss,
                "Metabolic Rate" = meta) %>% 
  apaSupp::tab_desc(caption = "Summary Statistics") %>% 
  flextable::footnote(i = 1:3, j = 1, 
                      value = as_paragraph(c("Weekly average of hours/day",
                                             "Daily average of 100s of calories above recommendation",
                                             "Weekly average of 100s of grams/week")))
```

### Correlation

```{r}
df_loss %>% 
  dplyr::select("Exercise Frequency" = exer,
                "Food Intake" = diet,
                "Weight Loss" = loss) %>% 
  apaSupp::tab_cor(caption = "Correlations")
```


### Visualizations

```{r, fig.cap= "D&H Figure 3.1 (page 45) An example with positive simple association and negative partial association - BASIC"}
df_loss %>% 
  ggplot(aes(x = diet,
             y = loss)) +
  geom_point(aes(shape = factor(exer))) 
```

```{r, fig.cap= "D&H Figure 3.1 (page 45) An example with positive simple association and negative partial association - BETTTER"}
df_loss %>% 
  ggplot(aes(x = diet,
             y = loss)) +
  geom_point(aes(shape = factor(exer),
                 color = factor(exer)),
             size = 3)  +
  theme_bw() +
  labs(x = "Observed Food Intake\nDaily average of 100s of calories above recommendation",
       y = "Observed Weight Loss\nWeekly average of 100s of grams/week",
       shape = "Exercise Frequency\nWeekly average of hours/day",
       color = "Exercise Frequency\nWeekly average of hours/day") +
  theme(legend.position = "bottom")
```

```{r, fig.cap= "D&H Figure 3.1 (page 45) An example with positive simple association and negative partial association - WORST"}
df_loss %>% 
  dplyr::mutate(exer = factor(exer)) %>% 
  ggplot(aes(x = diet,
             y = loss)) +
  geom_point(size = 3)  +
  theme_bw() +
  labs(x = "Observed Food Intake\n(Daily average of 100s of calories above recommendation)",
       y = "Observed Weight Loss\n(Weekly average of 100s of grams/week)",
       shape = "Exercise Frequency\n(Weekly average of hours/day)",
       color = "Exercise Frequency\n(Weekly average of hours/day)") +
  theme(legend.position = "bottom") +
  geom_smooth(aes(group = 1),
              method = "lm",
              formula = y ~ x,
              se = FALSE)
```


```{r, fig.cap= "D&H Figure 3.1 (page 45) An example with positive simple association and negative partial association - BEST"}
df_loss %>% 
  dplyr::mutate(exer = factor(exer)) %>% 
  ggplot(aes(x = diet,
             y = loss,
             group = exer)) +
  geom_point(aes(shape = exer,
                 color = exer),
             size = 3)  +
  theme_bw() +
  labs(x = "Observed Food Intake\n(Daily average of 100s of calories above recommendation)",
       y = "Observed Weight Loss\n(Weekly average of 100s of grams/week)",
       shape = "Exercise Frequency\n(Weekly average of hours/day)",
       color = "Exercise Frequency\n(Weekly average of hours/day)") +
  theme(legend.position = "bottom") +
  geom_smooth(aes(color = exer),
              method = "lm",
              formula = y ~ x,
              se = FALSE)
```



## REGRESSION ANALYSIS

### Fit the models

* The dependent variable (DV) is weight loss ($Y$)
* The independent variables (IVs) are exercise and diet ($X$)

```{r}
fit_lm_exer <- lm(loss ~ exer,
                  data = df_loss)

fit_lm_diet <- lm(loss ~ diet,
                  data = df_loss)

fit_lm_both <- lm(loss ~ exer + diet,
                  data = df_loss)

fit_lm_three<- lm(loss ~ exer + diet + meta,
                  data = df_loss)
```

### Output

```{r}
olsrr::ols_regress(fit_lm_exer)
```


```{r}
olsrr::ols_regress(fit_lm_diet)
```


```{r}
olsrr::ols_regress(fit_lm_both)
```



```{r}
olsrr::ols_regress(fit_lm_three)
```


### Equations

$$
\widehat{loss} = 4 + 1.75(exercise) \\

\widehat{loss} = 7.14 + 0.07(food) \\

\widehat{loss} = 6 + 2(exercise) -0.5(food) \\

\widehat{loss} = -1.636 + 1.045(exercise) -1.136(food) + 0.636(metabolism)\\
$$



### Tables

```{r}
list("Just\nExercise" = fit_lm_exer,
     "Just\nDiet" = fit_lm_diet,
     "Exercise\nand Diet" = fit_lm_both,
     "Add\nMetabolism" = fit_lm_three) %>% 
  modelsummary::modelsummary(output = "flextable",
                             fmt = 2,
                             estimate = "{estimate}",
                             statistic = NULL,
                             gof_map = NA,
                             coef_rename = c("exer" = "Exercise Frequency",
                                             "diet" = "Food Intake",
                                             "meta" = "Metabolism")) %>% 
  apaSupp::theme_apa(caption = "Parameter Estimates for Weight Loss Regressed on Exercise Frequency and Food Intake - BASIC")
```


```{r}
tab_mods <- list("Just Exercise" = fit_lm_exer,
                 "Just Diet" = fit_lm_diet,
                 "Exercise\nand Diet" = fit_lm_both,
     "Add\nMetabolism" = fit_lm_three) %>%  
  modelsummary::modelsummary(output = "flextable",
                             fmt = 2,
                             estimate = "{estimate} ({std.error})",
                             statistic = "{p.value}{stars}",
                             stars = c('*' = .05, '**' = .01, '***' = .001),
                             gof_map = c("r.squared", 
                                         #"adj.r.squared",
                                         "rmse"),
                             coef_rename = c("exer" = "Exercise Frequency",
                                             "diet" = "Food Intake",
                                             "meta" = "Metabolism")) %>% 
  apaSupp::theme_apa(caption = "Parameter Estimates for Weight Loss Regressed on Exercise Frequency and Food Intake - BETTER",
                     p_note = "apa",
                     general_note = "Dependent variable is average weekly weight lost in 100s of grams. ") %>% 
  flextable::hline(i = 8) %>% 
  flextable::footnote(i = c(3, 5, 7), j = 1, 
                      value = as_paragraph(c("Weekly average of hours/day",
                                             "Rate",
                                             "Daily average of 100s of calories above recommendation")))

tab_mods
```

```{r}
flextable::save_as_docx(tab_mods,
                        path = "tab_mods.docx")
```




```{r}
modelsummary::modelsummary(models = list("Weight Loss" = fit_lm_both),
                           output = "flextable",
                           fmt = 2,
                           coef_rename = c("exer" = "Exercise Frequency",
                                           "diet" = "Food Intake")) %>% 
  apaSupp::theme_apa(caption = "Parameter Estimates for Weight Loss Regressed on Exercise Frequency and Food Intake - BETTER",
                     p_note = "apa",
                     general_note = "Dependent variable is average weekly weight lost in 100s of grams. ") %>% 
  flextable::hline(i = 6) %>% 
  flextable::footnote(i = c(3, 5), j = 1, 
                      value = as_paragraph(c("Weekly average of hours/day",
                                             "Daily average of 100s of calories above recommendation")))
```





### Visualization

```{r}
eq <- function(x, y){6 + 2*x -.5 *y}

xlims <- 0:4
ylims <- 0:8

persp(x = xlims,
        y = ylims,
        z = outer(xlims, ylims, eq),
      xlab = "Exercise",
      ylab = "Food Consumed",
      zlab = "Weight Loss",
      shade = .3,
      theta = 200, 
      phi = 15, 
      r = 5,
      d = .5,
      ticktype = 'detailed')
```



```{r}
interactions::interact_plot(model = fit_lm_both,
                            pred = diet,
                            modx = exer,
                            modx.values = 0:4,
                            x.label = "Food Consumption, 100s calories above 1000",
                            y.label = "Weight Loss, 100s of grams",
                            legend.main = "Exercise,\nhours/day") +
  theme_bw() 
```



```{r}
interactions::interact_plot(model = fit_lm_both,
                            pred = exer,
                            modx = diet,
                            modx.values = c(2, 4, 6, 8),
                            x.label = "Daily Exercise, hours",
                            y.label = "Weight Loss, 100s of grams",
                            legend.main = "Food Consumption\n100s calories") +
  theme_bw() 
```


```{r}
interactions::interact_plot(model = fit_lm_three,
                            pred = exer,
                            modx = diet,
                            mod2 = meta,
                            modx.values = c(8, 5, 2),
                            modx.labels = c("1800 cal",
                                            "1500 cal",
                                            "1200 cal"),
                            mod2.values = c(16, 22.75),
                            mod2.labels = c("Low Metabolic Rate\n(Q1 = 16.00)", 
                                            "High Metabolic Rate\n(Q3 = 22.75)"),
                            x.label = "Daily Exercise, hours",
                            y.label = "Weight Loss, grams/week",
                            legend.main = "Food Consumption") +
  theme_bw() +
  theme(legend.position = c(0, 1),
        legend.justification = c(-0.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  scale_y_continuous(breaks = c(0, 5, 10, 15),
                     labels = c(0, 500, 1000, 1500))
```






## EFFECT SIZES

### Semipartial Correlation 

Unique contribution


```{r}
effectsize::r2_semipartial(fit_lm_both)
```

```{r}
effectsize::r2_semipartial(fit_lm_three)
```




### Eta Squared


```{r}
effectsize::eta_squared(fit_lm_both, partial = FALSE)
```

```{r}
effectsize::eta_squared(fit_lm_three, partial = FALSE)
```

```{r}
effectsize::eta_squared(fit_lm_both, partial = TRUE)
```

```{r}
effectsize::eta_squared(fit_lm_three, partial = TRUE)
```


### Standardized Regression Coefficients

```{r}
parameters::standardise_parameters(fit_lm_both)
```



```{r}
parameters::standardise_parameters(fit_lm_three)
```




## VARIANCE INFLATION FACTORS

```{r}
car::vif(fit_lm_both)
```

```{r}
car::vif(fit_lm_three)
```


## CONCLUSION: SPECIFIC QUESTIONS


```{r}
coef(fit_lm_both)
```


```{r}
confint(fit_lm_both)
```                               



$$
\widehat{loss} = 6 + 2(exercise) -0.5(food)
$$

Suppose an examination of the literature on diet, exercise, and weight loss led you to the following conclusions: 

1. People who eat the **minimum number of calories** to maintain good health and who **do not exercise** at all will lose how much?


```{r}
fit_lm_both %>% 
  emmeans::emmeans(~ exer + diet,
                   at = list(exer = 0,
                             diet = 0))
  
```

> The estimated marginal mean weight loss is 600 grams per week among thoes who eat the minimum number of calories to maintain good health and who do not exercise at, *b* = 6.00, 95% *CI* = [2.99, 9.01].



2. If food intake is held constant, then **each 1 hour of average daily exercise** leads to how much more weight loss?

```{r}
fit_lm_both %>% 
  emmeans::emmeans(~ exer + diet,
                   at = list(exer = 0:4))
  
```


> Holding food consumption constant, each additional hour of daily exercise is associated with 200 grams additional weight lost per week, *b* = 2.00, 95% *CI* = [1.21, 2.79].




3. If exercise is held constant, then **each one unit of food intake per day** (i.e., 100 calories) above the minimum to maintain good health translates to much weight gain?

```{r}
fit_lm_both %>% 
  emmeans::emmeans(~ diet,
                   at = list(diet = 0:4))
  
```

> Holding exercise constant, each additional 100 calories of food consumed daily is associated with 50 grams weight gained per week, *b* = -0.50, 95% *CI* = [1.21, 2.79]. 



4. If a person exercises 2 hours per week and eats 600 more calories than the minimum, how much weight would they loose weekly on average?


```{r}
fit_lm_both %>% 
  emmeans::emmeans(~ exer + diet,
                   at = list(exer = 2,
                             diet = 6))
```


> On average, 700 grams were lost per week amonth participants who exrcised 2 hours a week and consumed 6,600 more calories, 95% *CI* [572, 828]. 





