# Multiple Regression: "Sport Skill"


Darlington & Hayes, Chapter 3's second example


```{r}
# library(remotes)
# remotes::install_github("sarbearschwartz/apaSupp")

library(tidyverse)
library(flextable)
library(apaSupp)
library(modelsummary)
library(parameters)
library(effectsize)
library(interactions)
library(performance)
library(effects)
library(ggResidpanel)
```




## PURPOSE

### Research Question

> Are sports skills related to preference?

### Data Description

Independent Variables (IV, $X$)
* `soft` skill at softball
* `basket` skill at basketball

Dependent Variable (DV, $Y$)
* `pref` On a scale from 1 to 9, which sport do you prefer?”
    + 1 = much prefer softball
    + 5 = no preference
    + 9 = much prefer basketball


```{r}
df_sport <- tibble::tribble(~id, ~soft, ~basket, ~pref,
                             1,   4, 17, 1,
                             2,  56, 60, 3,
                             3,  25,  3, 8,
                             4,  50, 52, 4,
                             5,   5, 16, 2,
                             6,  72, 84, 2,
                             7, 100, 95, 5,
                             8,  39, 20, 8,
                             9,  81, 75, 5,
                            10,  61, 47, 7)
```

```{r}
df_sport %>% 
  dplyr::select("ID" = id,
                "Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H TAble 3.5 (pg 78) Skill at Softball, Basketball, and Preference") %>% 
  flextable::colformat_double(digits = 0)
```


## EXPLORATORY DATA ANALYSIS

### Univariate Statistics

Center: mean and median
Spread: standard deviation, range (max - min), interquartile range (Q3 - Q1)

```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  apaSupp::tab_desc(caption = "Summary Statistics")
```



### Univariate Visualizations


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = soft)
```


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = basket)
```


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = pref)
```




### Bivariate Statistics

```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  apaSupp::tab_cor(caption = "Unadjusted, Pairwise Correlations")
```




### Bivariate Visualization




```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  data.frame %>% 
  GGally::ggscatmat() +
  theme_bw()
```



```{r}
df_sport %>% 
  ggplot(aes(x = soft,        
           y = pref)) +   
  geom_point() +                    
  geom_smooth(method = "lm",
              formula = y ~ x) +
  ggpubr::stat_regline_equation(label.x = 75,
                                label.y = 1,
                                size = 6) +
  theme_bw() +
  labs(x = "Softball Skill",
       y = "Preference")
```


```{r}
df_sport %>% 
  ggplot(aes(x = basket,        
             y = pref)) +   
  geom_point() +                    
  geom_smooth(method = "lm",
              formula = y ~ x) +
  ggpubr::stat_regline_equation(label.x = 75,
                                label.y = 1,
                                size = 6) +
  theme_bw() +
  labs(x = "Basketball Skill",
       y = "Preference")
```



## REGRESSION ANALYSIS

* The dependent variable (DV) is preference ($Y$)
* The independent variable (IVs) are skill levels ($X$)

### Fit the Models

```{r}
fit_lm_soft <- lm(pref ~ soft,
                   data = df_sport)  

fit_lm_basket <- lm(pref ~ basket,
                   data = df_sport)

fit_lm_both <- lm(pref ~ soft + basket,
                   data = df_sport)
```



```{r}
list("Just\nSoftball" = fit_lm_soft,
     "Just\nBasketball" = fit_lm_basket,
     "Softball\nand Basketball" = fit_lm_both) %>% 
  modelsummary::modelsummary(output = "flextable",
                             fmt = 2,
                             estimate = "{estimate} ({std.error})",
                             statistic = "{p.value}{stars}",
                             stars = c('*' = .05, '**' = .01, '***' = .001),
                             gof_map = c("r.squared", 
                                         #"adj.r.squared",
                                         "rmse"),
                             coef_rename = c("soft" = "Softball Skill",
                                             "basket" = "Basketball Skill")) %>% 
  apaSupp::theme_apa(caption = "Parameter Estimates for Preference REgressed on Skills in Softball and Basketball",
                     p_note = "apa") %>% 
  flextable::hline(i = 6) 
```


```{block type='rmdimportant', echo=TRUE}
We call a set of regressors **complementary** if $R^2$ for the set exceeds the sum of the individual values of $r^2_{YX}$ . Thus, complementarity and **collinearity** are opposites, though either can occur only when regressors in a set are **intercorrelated**.
```




### Semipartial Correlation



```{r}
effectsize::r2_semipartial(fit_lm_both)
```


### Venn Diagram - Variances 

https://freetools.touchpoint.com/venn-diagram-template-generator 

```{r}
ABCD <- var(df_sport$pref)
ABCD
```

```{r}
AB <- var(fit_lm_soft$residuals)
AB
```


```{r}
BC <- var(fit_lm_basket$residuals)
BC
```


```{r}
D <- var(fit_lm_both$residuals)
D
```


```{r}
ABC <- ABCD - D
ABC
```


```{r}
A <- ABC - BC
A
```

```{r}
C <- ABC - AB
C
```

```{r}
B <- AB + BC - ABC
B
```

### Venn Diagram - Proportions

Total Variance in Preference Explained by Both Skills

```{r}
R2 <- ABC/ABCD
R2
```

Variance in Preference Uniquely Explained by Softball skills, across all Basketball skill levels

```{r}
sr2_soft <- A/ABCD
sr2_soft
```


Variance in Preference Uniquely Explained by Basketball skills, across all Softball skill levels

```{r}
sr2_basket <- C/ABCD
sr2_basket
```

Variance in Preference Explained by Softball skills, when holding Basketball skill constant

```{r}
pr2_soft <- A/(A + D)
pr2_soft
```



```{r}
pr2_basket <- C/(C + D)
pr2_basket
```

### Residual Diagnostics

```{r}
performance::check_residuals(fit_lm_both)
```

```{r}
ggResidpanel::resid_panel(fit_lm_both)
```
