# D&H Ch3b - Multiple Regression: "Sport Skill"

Compiled: `r format(Sys.time(), '%B %d, %Y')`

Darlington & Hayes, Chapter 3's second example


```{r, error=FALSE, warning=FALSE, message=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")


library(tidyverse)
library(apaSupp)     
library(naniar)
library(GGally)
library(corrplot)    
library(ggpubr)  
library(interactions)
library(effectsize)
library(performance)
library(ggResidpanel)
```




## PURPOSE

### Research Question

> Are sports skills related to preference?

### Data Description


#### Variables

Independent Variables (IV, $X$)
* `soft` skill at softball
* `basket` skill at basketball

Dependent Variable (DV, $Y$)
* `pref` On a scale from 1 to 9, which sport do you prefer?”
+ 1 = much prefer softball
+ 5 = no preference
+ 9 = much prefer basketball


```{r}
df_sport <- tibble::tribble(~id, ~soft, ~basket, ~pref,
                            1,   4, 17, 1,
                            2,  56, 60, 3,
                            3,  25,  3, 8,
                            4,  50, 52, 4,
                            5,   5, 16, 2,
                            6,  72, 84, 2,
                            7, 100, 95, 5,
                            8,  39, 20, 8,
                            9,  81, 75, 5,
                            10,  61, 47, 7)
```

```{r}
df_sport %>% 
  dplyr::select("ID" = id,
                "Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "D&H TAble 3.5 (pg 78) Skill at Softball, Basketball, and Preference") %>% 
  flextable::colformat_double(digits = 0)
```


## EXPLORATORY DATA ANALYSIS

### Univariate Statistics

Center: mean and median
Spread: standard deviation, range (max - min), interquartile range (Q3 - Q1)

```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  apaSupp::tab_desc(caption = "Summary Statistics")
```



### Univariate Visualizations


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = soft)
```


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = basket)
```


```{r}
df_sport %>% 
  apaSupp::spicy_histos(var = pref)
```




### Bivariate Statistics

```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  apaSupp::tab_cor(caption = "Unadjusted, Pairwise Correlations")
```




### Bivariate Visualization




```{r}
df_sport %>% 
  dplyr::select("Softball Skill" = soft,
                "Basketball Skill" = basket,
                "Preference Score" = pref) %>%
  data.frame %>% 
  GGally::ggscatmat() +
  theme_bw()
```



```{r}
df_sport %>% 
  ggplot(aes(x = soft,        
             y = pref)) +   
  geom_point() +                    
  geom_smooth(method = "lm",
              formula = y ~ x) +
  ggpubr::stat_regline_equation(label.x = 75,
                                label.y = 1,
                                size = 6) +
  theme_bw() +
  labs(x = "Softball Skill",
       y = "Preference")
```


```{r}
df_sport %>% 
  ggplot(aes(x = basket,        
             y = pref)) +   
  geom_point() +                    
  geom_smooth(method = "lm",
              formula = y ~ x) +
  ggpubr::stat_regline_equation(label.x = 75,
                                label.y = 1,
                                size = 6) +
  theme_bw() +
  labs(x = "Basketball Skill",
       y = "Preference")
```



## REGRESSION ANALYSIS

* The dependent variable (DV) is preference ($Y$)
* The independent variable (IVs) are skill levels ($X$)

### Fit the Models

```{r}
fit_lm_soft <- lm(pref ~ soft,
                  data = df_sport)  

fit_lm_basket <- lm(pref ~ basket,
                    data = df_sport)

fit_lm_both <- lm(pref ~ soft + basket,
                  data = df_sport)
```



```{r}
tab_lm3 <- apaSupp::tab_lms(list("Softball"   = fit_lm_soft, 
                                 "Basketball" = fit_lm_basket, 
                                 "Both"       = fit_lm_both),
                            var_labels = c("soft" = "Softball Skill",
                                           "basket" = "Basketball Skill"),
                            caption = "Parameter Estimates for Sport Preference Regression on Skill Level in Softball and Basketball",
                            general_note = "Dependent variable is preference rating on a scale of 1 (much prefer softball) to  9 (much prefer basketball).  Both softball and Baksetball skill levels are on a scale of 0 to 100.",
                            narrow = TRUE,
                            d = 3) 

tab_lm3
```


```{block type='rmdimportant', echo=TRUE}
We call a set of regressors **complementary** if $R^2$ for the set exceeds the sum of the individual values of $r^2_{YX}$ . Thus, complementarity and **collinearity** are opposites, though either can occur only when regressors in a set are **intercorrelated**.
```


### Visualize

```{r}
interactions::interact_plot(model = fit_lm_both,
                            pred = soft,
                            modx = basket,
                            modx.values = c(25, 50, 75),
                            legend.main = "Basketball Skill\n(0-100)",
                            interval = TRUE) +
  theme_bw() +
  labs(x = "Softball Skill, (0-100)",
       y = "Estimated Marginal Mean Preference\n1 = much prefer softball\n9 = much prefer basketball") +
  coord_cartesian(ylim = c(1, 9)) +
  scale_y_continuous(breaks = 1:9) +
  theme(legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```




### Semipartial Correlation

```{r}
apaSupp::tab_lm(fit_lm_both)
```



```{r}
apaSupp::tab_lm(fit_lm_both)
```


```{r}
effectsize::r2_semipartial(fit_lm_both)
```




### Venn Diagram - Variances 

https://freetools.touchpoint.com/venn-diagram-template-generator 

```{r}
ABCD <- var(df_sport$pref)
ABCD
```

```{r}
AB <- var(fit_lm_soft$residuals)
AB
```


```{r}
BC <- var(fit_lm_basket$residuals)
BC
```


```{r}
D <- var(fit_lm_both$residuals)
D
```


```{r}
ABC <- ABCD - D
ABC
```


```{r}
A <- ABC - BC
A
```

```{r}
C <- ABC - AB
C
```

```{r}
B <- AB + BC - ABC
B
```

### Venn Diagram - Proportions

Total Variance in Preference Explained by Both Skills

```{r}
R2 <- ABC/ABCD
R2
```

Variance in Preference Uniquely Explained by Softball skills, across all Basketball skill levels

```{r}
sr2_soft <- A/ABCD
sr2_soft
```


Variance in Preference Uniquely Explained by Basketball skills, across all Softball skill levels

```{r}
sr2_basket <- C/ABCD
sr2_basket
```

Variance in Preference Explained by Softball skills, when holding Basketball skill constant

```{r}
pr2_soft <- A/(A + D)
pr2_soft
```



```{r}
pr2_basket <- C/(C + D)
pr2_basket
```

### Residual Diagnostics

```{r}
performance::check_residuals(fit_lm_both)
```

```{r}
ggResidpanel::resid_panel(fit_lm_both)
```
