# D&H Ch8 - Regressor Importance: "politics"


Darlington & Hayes, Chapter 8's  example


```{r, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")

library(tidyverse) 
library(haven)
library(flextable)
library(apaSupp)
library(car)
library(rempsyc)
library(parameters)
library(performance)
library(interactions)
library(ggResidpanel)
library(relaimpo)
library(dominanceanalysis)
library(domir)
```

```{r}
flextable::set_flextable_defaults(digits = 2)
```



## PURPOSE


```{block type='rmdlightbulb', echo=TRUE}
RESEARCH QUESTION:

RQ1) Is news consumption via national news broadcast and newspaper correlated with knowledge of the political process, after accounting for sex and age?

RQ2) If so, do both sources of news have the same effect on political knowledge?

RQ3) Is listening to political talk radio more or less important than watching the national network news broadcast?

```



### Data Import

```{block type='rmddownload', echo=TRUE}
You can download the `politics` dataset here:

* [SPSS format (.sav)](https://raw.githubusercontent.com/SarBearSchwartz/v4_regression/master/data/darlington_hayes/politics.sav)
```



```{r, echo=FALSE}
df_spss <- haven::read_sav("data/darlington_hayes/politics.sav")
```

```{r, eval=FALSE}
df_spss <- haven::read_sav("politics.sav")
```


```{r}
tibble::glimpse(df_spss)
```

```{r}
summary(df_spss)
```


```{block type='rmdimportant', echo=TRUE}
When importing data from SPSS (.sav), you need to be careful how categorical vairables are stored.  
```



```{r}
df_fac <- df_spss %>% 
  haven::as_factor()
```


```{r}
tibble::glimpse(df_fac)
```

```{r}
summary(df_fac)
```



### Data Description


National survey of residents of the United States.

Participants were asked a set of questios used to quantify their knowledge of politics, politicians, and the political process.

Assumption: knowledge is caused by exposure to information  

#### Variables

**Dependent Variable** (outcome, Y)

* `pknow` knowledge of the political process


**Independent Variables** (predictors or regressors, X's)  

Frequency of Exposure (days per week)...

* `talkrad` listening to political talk radio
* `natnews` watch national news broadcasts
* `npnews`  read newspaper
* `locnews` watch local news

Covariates

* `age` age in years, 18-90
* `sex` sex, Male vs. Female


```{block type='rmdimportant', echo=TRUE}
Categorical variables MUST be declare as FACTORS and the FIRST level listed is treated as the reference category.  
```


```{r}
df_pol <- df_spss %>% 
  haven::as_factor() %>% 
  haven::zap_label() %>% 
  tibble::rowid_to_column(var = "id") %>% 
  dplyr::mutate(news_sum = (natnews + npnews)/2) %>% 
  dplyr::mutate(news_dif = (natnews - npnews)/2)
```


```{r}
tibble::glimpse(df_pol)
```

```{r}
summary(df_pol)
```

## REGRESSION ANALYSIS

### RQ1) Fit Main Model

```{r}
fit_pol_1 <- lm(pknow ~ natnews + npnews + age + sex,
                data = df_pol)
```

```{r}
apaSupp::tab_lm(fit_pol_1,
                var_labels = c(natnews = "National News",
                               npnews = "Newspapers",
                               age = "Age",
                               sex = "Sex"),
                d = 3) %>% 
  flextable::width(j = 1, width = 1.5)
```


### RQ2) Parameter Equivalence

```{r}
fit_pol_2 <- lm(pknow ~ news_sum + age + sex,
                data = df_pol)

fit_pol_3 <- lm(pknow ~ news_sum + news_dif + age + sex,
                data = df_pol)
```

```{r}
apaSupp::tab_lms(list(fit_pol_1, fit_pol_2, fit_pol_3),
                 var_labels = c(news_sum = "News, sum",
                                news_dif = "News, dif",
                                age = "Age",
                                sex = "Sex"))
```
```{r}
car::linearHypothesis(fit_pol_1, c("natnews - npnews"))
```

### RQ3) Variable Contribution

#### Variable Importance


```{r}
lm(pknow ~ npnews + locnews + talkrad + natnews,
   data = df_pol) %>% 
  caret::varImp()
```


```{r}
lm(pknow ~ npnews + locnews + talkrad + natnews,
   data = df_pol) %>% 
  relaimpo::calc.relimp(type = "lmg", importance = TRUE)
```


#### Relative Improvement in Fit

Set A: 
* newspaper
* localnews

Add:
* national news
* talk ratio

```{r}
data.frame(A = c("None",
                 "Newspaper",
                 "Local News",
                 "Both"),
           base = c("pknow ~ 1", 
                    "pknow ~ npnews", 
                    "pknow ~ locnews", 
                    "pknow ~ npnews+locnews")) %>% 
  dplyr::mutate(add_talk = paste0(base, "+talkrad")) %>% 
  dplyr::mutate(add_natn = paste0(base, "+natnews")) %>% 
  dplyr::mutate(fit_base = purrr::map(base,
                                      ~lm(.x, data = df_pol))) %>%
  dplyr::mutate(fit_talk = purrr::map(add_talk,
                                      ~lm(.x, data = df_pol))) %>%
  dplyr::mutate(fit_natn = purrr::map(add_natn,
                                      ~lm(.x, data = df_pol))) %>%
  dplyr::mutate(R2 = purrr::map_dbl(fit_base,
                                    ~ broom::glance(.x)$r.squared)) %>% 
  dplyr::mutate(R2i = purrr::map_dbl(fit_talk,
                                     ~ broom::glance(.x)$r.squared)) %>% 
  dplyr::mutate(R2j = purrr::map_dbl(fit_natn,
                                     ~ broom::glance(.x)$r.squared)) %>% 
  dplyr::mutate(R  = sqrt(R2)) %>% 
  dplyr::mutate(Ri = sqrt(R2i)) %>% 
  dplyr::mutate(Rj = sqrt(R2j)) %>% 
  dplyr::mutate(dRi = Ri - R) %>% 
  dplyr::mutate(dRj = Rj - R) %>% 
  dplyr::select("Set A subset" = A,
                "R" = R, 
                "Adding\nTalk Radio\nR" = Ri,
                "Adding\nNational News\nR" = Rj,
                "Talk Radio\nChange\nin R" = Ri,
                "National News\nChange\nin R" = Rj) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Table 8.2 - Relative Improvement in Fit for Dominance Computations",
                     d = 3)
```



#### Dominance Ananlysis

see: https://cran.r-project.org/web/packages/domir/vignettes/domir_basics.html



```{r}
domir(
  pknow ~ npnews + locnews + talkrad + natnews, 
  function(formula) {
    lm_model <- lm(formula, data = df_pol)
    summary(lm_model)[["r.squared"]]
  }
)
```





