# D&H Ch9 - Multicategorical Regressors: "married"

Compiled: `r format(Sys.time(), '%B %d, %Y')`

Darlington & Hayes, Chapter 9's  example


```{r, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")

library(tidyverse) 
library(haven)
library(flextable)
library(apaSupp)
library(car)
library(rempsyc)
library(parameters)
library(performance)
library(interactions)
library(ggResidpanel)
```





## PURPOSE


```{block type='rmdlightbulb', echo=TRUE}
RESEARCH QUESTION:

RQ 1) Does life satisfaction differ by marital status?

RQ2) Is there an association between income and life satisfaction, after controlling for marital status?

```



### Data Import

```{block type='rmddownload', echo=TRUE}
You can download the `married` dataset here:

* [SPSS format (.sav)](https://raw.githubusercontent.com/SarBearSchwartz/v4_regression/master/data/darlington_hayes/married.sav)
```



```{r, echo=FALSE}
df_spss <- haven::read_sav("data/darlington_hayes/married.sav")
```

```{r, eval=FALSE}
df_spss <- haven::read_sav("married.sav")
```


```{r}
tibble::glimpse(df_spss)
```

```{r}
summary(df_spss)
```


```{block type='rmdimportant', echo=TRUE}
When importing data from SPSS (.sav), you need to be careful how categorical vairables are stored.  
```



```{r}
df_fac <- df_spss %>% 
  haven::as_factor()
```


```{r}
tibble::glimpse(df_fac)
```

```{r}
summary(df_fac)
```



### Data Description

#### Variables

Dependent Variable (outcome, Y)

* `statis` life satisfaction, scale 1-100 (composite of many items)

Independent Variables (predictors or regressors, X's)  

* `mstatus` marital status: 1 = married, 2 = divorced, 3 = single, 4 = widowed  
* `income` annual income, $1000s/year  
* `sex` 0 = female, 1 = male  



```{block type='rmdimportant', echo=TRUE}
Categorical variables MUST be declare as FACTORS and the FIRST level listed is treated as the reference category.  
```


```{r}
df_sat <- df_spss %>% 
  haven::as_factor() %>% 
  tibble::rowid_to_column(var = "id") %>% 
  dplyr::mutate(mstatus = forcats::fct_recode(mstatus,
                                              "Married"  = "married",
                                              "Divorced" = "divorced",
                                              "Single"   = "single",
                                              "Widowed"  = "widowed")) %>% 
  dplyr::mutate(sex = forcats::fct_recode(sex,
                                          "Female" = "female",
                                          "Male" = "male")) %>% 
  dplyr::mutate(mstat_mar = forcats::fct_relevel(mstatus, "Married",  after = 0)) %>%
  dplyr::mutate(mstat_div = forcats::fct_relevel(mstatus, "Divorced", after = 0)) %>% 
  dplyr::mutate(mstat_sin = forcats::fct_relevel(mstatus, "Single",   after = 0)) %>% 
  dplyr::mutate(mstat_wid = forcats::fct_relevel(mstatus, "Widowed",  after = 0)) %>% 
  dplyr::mutate(male   = forcats::fct_relevel(sex, "Male",   after = 1)) %>% 
  dplyr::mutate(female = forcats::fct_relevel(sex, "Female", after = 1))
```


```{r}
tibble::glimpse(df_sat)
```

```{r}
summary(df_sat)
```



## EXPLORATORY DATA ANALYSIS

```{block type='rmdimportant', echo=TRUE}
Before embarking on any inferencial anlaysis or modeling, always get familiar with your variables one at a time *(univariate)*, as well as pairwise *(bivariate)*.
```

### Summary Statistics



#### Univariate

```{r}
df_sat %>% 
  dplyr::select("Sex" = sex,
                "Marital Status" = mstatus) %>% 
  apaSupp::tab_freq(caption = "Description of Categorical Measures")
```

```{r}
df_sat %>% 
  dplyr::select("Annual Income" = income,
                "Life Satisfaction" = satis) %>% 
  apaSupp::tab_desc(caption = "Description of Continuous Measures",
                    general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.")
```




#### Bivariate

```{r}
df_sat %>% 
  dplyr::select(mstatus,
                "Sex" = sex,
                "Annual Income" = income,
                "Life Satisfaction" = satis) %>% 
  apaSupp::table1_apa(split = mstatus,
                      caption = "Description of Continuous Measures",
                      general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.")
```



```{r}
cor.test(~ satis + income, data = df_sat)
```






### Visualize Distributions

#### Univariate 


```{block type='rmdimportant', echo=TRUE}
Always plot your data first!
```

```{r}
df_sat %>% 
  apaSupp::spicy_histos(var = satis,
                        lab = "Life Satisfacton")
```

```{r}
df_sat %>% 
  apaSupp::spicy_histos(var = income,
                        lab = "Annual Income, $1000/yr")
```



#### Bivariate


```{r, fig.width=6, fig.height=6}
df_sat %>% 
  dplyr::select("Life Satisfaction" = satis,
                "Income" = income,
                "Marital Satus" = mstatus,
                "Sex" = sex) %>%
  data.frame %>% 
  GGally::ggpairs() +
  theme_bw()
```





```{r, fig.width=6, fig.height=8}
df_sat %>% 
  apaSupp::spicy_histos(var = satis,
                        split = mstatus,
                        lab = "Life Satisfacton")
```


```{r}
df_sat %>% 
  ggplot(aes(x = mstatus,
             y = satis,
             group = mstatus)) +
  geom_violin(fill = "gray") +
  geom_boxplot(width = .25) +
  geom_point(color = "red",
             position = position_jitter(width = .1),
             size = 2) +
  theme_bw() +
  labs(x = "Marital Status",
       y = "Observed Life Satisfaction")
```



```{r, fig.width=6, fig.height=5}
df_sat %>% 
  apaSupp::spicy_histos(var = satis,
                        split = sex,
                        lab = "Life Satisfacton")
```


```{r}
df_sat %>% 
  ggplot(aes(x = sex,
             y = satis,
             group = sex)) +
  geom_violin(fill = "gray") +
  geom_boxplot(width = .25) +
  geom_point(color = "red",
             position = position_jitter(width = .1),
             size = 2) +
  theme_bw() +
  labs(x = NULL,
       y = "Observed Life Satisfaction")
```


```{r}
df_sat %>% 
  ggplot(aes(x = income,
             y = satis)) +
  geom_point(aes(color = mstatus,
                 shape = mstatus),
             size = 2) +
  theme_bw() +
  geom_smooth(method = "lm",
              formula = y ~ x,
              color = "gray",
              alpha = .2) +
  labs(x = "Annual Income, $1000/yr",
       y = "Observed Life Satisfaction, 1-100",
       color = NULL,
       shape = NULL) +
  theme(legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.background = element_rect(color = "black"))
```



```{r}
df_sat %>% 
  ggplot(aes(x = income,
             y = satis)) +
  geom_point() +
  theme_bw() +
  geom_smooth(method = "lm",
              formula = y ~ x,
              alpha = .2) +
  labs(x = "Annual Income, $1000/yr",
       y = "Observed Life Satisfaction") +
  facet_wrap(~ mstatus)
```


```{r}
df_sat %>% 
  ggplot(aes(x = income,
             y = satis)) +
  geom_point(aes(color = mstatus,
                 shape = mstatus),
             size = 2) +
  theme_bw() +
  geom_smooth(aes(color = mstatus),
              method = "lm",
              formula = y ~ x,
              se = FALSE) +
  labs(x = "Annual Income, $1000/yr",
       y = "Observed Life Satisfaction, 1-100",
       color = NULL,
       shape = NULL) +
  theme(legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.background = element_rect(color = "black"))
```





```{r}
df_sat %>% 
  ggplot(aes(x = income,
             y = satis)) +
  geom_point(aes(color = sex,
                 shape = sex),
             size = 2) +
  theme_bw() +
  geom_smooth(aes(color = sex),
              method = "lm",
              formula = y ~ x,
              alpha = .2) +
  labs(x = "Annual Income, $1000/yr",
       y = "Observed Life Satisfaction, 1-100",
       color = NULL,
       shape = NULL) +
  theme(legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.background = element_rect(color = "black"))
```




## ANALYSIS OF VARIANE

* The dependent variable (DV) is Life Satisfaction ($satis$)

### By Marital Status

* The independent variable (IV) is marital status, with widowed as the reference category to match the textbook.

> NOTE: SPSS used the LAST listed category level as the reference group whereas R uses the FIRST.

```{r}
fit_aov_1 <- afex::aov_4(satis ~ mstat_wid + (1|id), 
                         data = df_sat)

summary(fit_aov_1)
```

```{r}
fit_aov_1 %>% 
  emmeans::emmeans(~ mstat_wid)
```

```{r}
fit_aov_1 %>% 
  emmeans::emmeans(~ mstat_wid) %>% 
  pairs()
```




### By Sex

* The independent variable (IV) is sex.


```{r}
fit_aov_2 <- afex::aov_4(satis ~ sex + (1|id), 
                         data = df_sat)

summary(fit_aov_2)
```






## REGRESSION ANALYSIS

* The dependent variable (DV) is Life Satisfaction ($satis$)

### Roll of Marital Status

```{r}
fit_lm_1 <- lm(satis ~ mstat_wid, data = df_sat)
```

```{r}
apaSupp::tab_lm(fit_lm_1, 
                var_labels = c(mstat_wid = "Marital Status"),
                caption = "Parameter Estimates for Life Satisfaction Regressed on Marital Status")
```




```{r}
anova(fit_lm_1)
```



```{r}
fit_lm_1 %>% 
  emmeans::emmeans(~ mstat_wid)
```

```{r}
fit_lm_1 %>% 
  emmeans::emmeans(~ mstat_wid) %>% 
  pairs()
```



#### Change the Reference Category

```{r}
fit_lm_1m <- lm(satis ~ mstat_mar, data = df_sat)
fit_lm_1d <- lm(satis ~ mstat_div, data = df_sat)
fit_lm_1s <- lm(satis ~ mstat_sin, data = df_sat)
fit_lm_1w <- lm(satis ~ mstat_wid, data = df_sat)
```


```{r}
apaSupp::tab_lms(list(fit_lm_1m, fit_lm_1d, fit_lm_1s, fit_lm_1w),
                 var_labels = c(mstat_mar = "Marital Status",
                                mstat_div = "Marital Status",
                                mstat_sin = "Marital Status",
                                mstat_wid = "Marital Status"),
                 caption = "Comparison of Regression Models Changing the Reference Category",
                 d = 1,
                 narrow = TRUE) %>% 
  flextable::autofit()
```




### Roll of Income

#### Unadjusted Model

No Covariates

```{r}
fit_lm_2 <- lm(satis ~ income, data = df_sat)

apaSupp::tab_lm(fit_lm_2)
```



#### Adjust for Marital Status

Covary Marital Status

```{r}
fit_lm_3 <- lm(satis ~ income + mstat_wid, data = df_sat)

apaSupp::tab_lm(fit_lm_3)
```




```{r}
apaSupp::tab_lms(list(fit_lm_2, fit_lm_3),
                 var_labels = c(mstat_wid = "Marital Status",
                                income  = "Annual Income"),
                 caption = "Comparison for Models Regressing Life Satisfaction on Annual Income, with and without Controlling for Marital Status",
                 general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.",
                 d = 3) %>% 
  flextable::autofit()
```


```{r}
apaSupp::tab_lm(fit_lm_2,
                var_labels = c(income    = "Annual Income"),
                caption = "Parameter Estimates for Life Satisfaction Regressed on Annual Income Ignoring Marital Status",
                general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.") %>% 
  flextable::autofit()
```

```{r}
apaSupp::tab_lm(fit_lm_3,
                var_labels = c(income    = "Annual Income",
                               mstat_wid = "Marital Status"),
                caption = "Parameter Estimates for Life Satisfaction Regressed on Annual Income while Controlling for Marital Status",
                general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.") %>% 
  flextable::autofit()
```


```{r}
fit_lm_3 %>% 
  emmeans::emmeans(~ mstat_wid | income,
                   at = list(income = c(50, 56.9))) 
```

```{r}
interactions::interact_plot(model = fit_lm_3,
                            pred = income,
                            modx = mstat_wid)
```



```{r}
interactions::interact_plot(model = fit_lm_3,
                            pred = income,
                            modx = mstat_wid,
                            legend.main = "Marital Status",
                            interval = TRUE) +
  labs(x = "Annual Income, $1000s/yr",
       y = "Estimated Marginal Mean\nLife Satisfaction") +
  theme_bw() +
  theme(legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.key.width = unit(1.5, "cm"),
        legend.background = element_rect(color = "black"))
```



#### Adjust for Sex & Marital Status

Covary Sex and Marital Status

```{r}
fit_lm_4 <- lm(satis ~ income + mstat_wid + sex, data = df_sat)

anova(fit_lm_4)
```


```{r}
apaSupp::tab_lm(fit_lm_4,
                var_labels = c(income    = "Annual Income",
                               mstat_wid = "Marital Status",
                               sex       = "Male vs. Female"),
                show_single_row = "sex",
                caption = "Parameter Estimates for Life Satisfaction Regressed on Annual Income while Controlling for Marital Status and Sex",
                general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.") %>% 
  flextable::autofit()
```




```{r}
apaSupp::tab_lms(list(fit_lm_2, fit_lm_3, fit_lm_4),
                 var_labels = c(mstat_wid = "Marital Status",
                                income    = "Annual Income",
                                sex       = "Sex"),
                 caption = "Comparison for Models Regressing Life Satisfaction on Annual Income, with and without Controlling for Marital Status and Sex",
                 general_note = "Annual income is in $1000s per year.  Life Satisfaction is a composite score on a scale 1-100.",
                 d = 2,
                 narrow = TRUE)
```


## WRITE UP

```{block type='rmdlightbulb', echo=TRUE}
RESEARCH QUESTION:

RQ 1) Does life satisfaction differ by marital status?

RQ2) Is there an association between income and life satisfaction, after controlling for marital status?

```

### Methods Section


### Results Section


### Table(s)


### Figure(s)
