# Ex: Basic - Cancer Experiment 


```{r, comment=FALSE, message=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")

library(magrittr)       
library(tidyverse)   
library(broom)     
library(naniar)
library(corrplot)   
library(GGally)
library(gtsummary)
library(apaSupp)
library(performance)
library(interactions)
library(effects)
library(emmeans)
library(car)
library(ggResidpanel)
library(modelsummary)
library(ppcor)
library(jtools)
library(olsrr)
library(DescTools)
library(effectsize)
library(ggpubr)
```

```{r}
emmeans::emm_options(opt.digits = FALSE)  # revert to optimal digits
```


## PURPOSE

### Research Question

> For this example page: Does weight vary with age?


### Data Description

The `Cancer` data set contains 4 repeated measure of oral condition for 25 cancer patients to see if an aloe treatment helps with oral condition (ulcers).


```{r}
cancer_raw <- haven::read_spss("https://raw.githubusercontent.com/CEHS-research/eBook_ANOVA/master/data/Cancer.sav")
```


```{r}
tibble::glimpse(cancer_raw)
```

```{r}
cancer_clean <- cancer_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(trt = factor(trt,
                             labels = c("Placebo", 
                                        "Aloe Juice"))) %>% 
  dplyr::mutate(stage = factor(stage))

tibble::glimpse(cancer_clean)
```



```{r}
psych::headTail(cancer_clean) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Sample of Cancer Dataset")
```




## EXPLORATORY DATA ANALYSIS


```{block type='rmdimportant', echo=TRUE}
Before embarking on any inferencial anlaysis or modeling, always get familiar with your variables one at a time *(univariate)*, as well as pairwise *(bivariate)*.
```



### Summary Statistics

#### Univariate 

Center: mean and median
Spread: standard deviation, range (max - min), interquartile range (Q3 - Q1)

```{r}
cancer_clean %>% 
  dplyr::select("Age" = age,
                "Weight" = weighin) %>% 
  apaSupp::tab_desc(caption = "Description of Participants")
```



#### Bivariate 

```{r}
cancer_clean %>% 
  cor.test(~ age + weighin,
           data = .)
```





### Visualize Distributions


#### Univariate 


```{r}
cancer_clean %>% 
  apaSupp::spicy_histos(var = age)
```


```{r}
cancer_clean %>% 
  apaSupp::spicy_histos(var = weighin)
```


#### Bivariate 


```{r}
cancer_clean %>% 
  ggplot(aes(x = age,
             y = weighin)) +
  geom_point() +
  geom_smooth(aes(color = "Linear"), 
              method = "lm", 
              formula = y ~ x,
              se = TRUE) +  
  geom_smooth(aes(color = "Loess"),  
              method = "loess", 
              se = FALSE) +     
  theme_bw() +
  labs(x = "Age, years",
       y = "Weight, pounds",
       color = "Smoother") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"))
```


## REGRESSION ANALYSIS



```{block type='rmdlightbulb', echo=TRUE}
The `lm()` function must be supplied with at least two options:

* a formula:  `Y ~ X`
* a dataset: `data = XXXXXXX`

When a model is fit and directly saved as a named object via the assignment opperator (`<-`), no output is produced.
```


* The dependent variable (DV) is weight at intake ($Y$)
* The independent variable (IV) is age at intake ($X$)

### Fit Model

```{r}
fit_lm <- lm(weighin ~ age,
             data = cancer_clean)
```


### Parameter Estimates

```{r}
summary(fit_lm)
```



### Prediction Equation

```{r}
coef(fit_lm)
```



$$
\hat{weight} = 220.69 - 0.71 \times (age)
$$

### Table

```{r}
apaSupp::tab_lm(fit_lm)
```



### Estimated Marginal Means

```{r}
fit_lm %>% 
  emmeans::emmeans(~ age)
```

```{r}
fit_lm %>% 
  emmeans::emmeans(~ age,
                   at = list(age = c(30, 60, 90))) 
```


```{r}
fit_lm %>% 
  emmeans::emmeans(~ age,
                   at = list(age = seq(from = 30, to = 90, by = 10))) %>% 
  data.frame() %>% 
  dplyr::mutate(ci = glue::glue("[{apaSupp::apa2(lower.CL)}, {apaSupp::apa2(upper.CL)}]")) %>% 
  dplyr::select("Age" = age,
                "Est. Marginal Mean" = emmean,
                SE,
                "95% CI" = ci) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Estimated Marginal Means for Weight Regression on Age") %>% 
  flextable::colformat_double(j = 1, digits = 0) %>% 
  flextable::align(align = "center")
```


```{r}
fit_lm %>% 
  effects::Effect(focal.predictors = c("age"))
```


```{r}
fit_lm %>% 
  effects::Effect(focal.predictors = c("age"),
                  xlevels = list(age = seq(from = 30, to = 90, by = 1))) %>% 
  data.frame() 
```




```{r, fig.cap = "Weight Regressed on Age"}
fit_lm %>% 
  effects::Effect(focal.predictors = c("age"),
                  xlevels = list(age = seq(from = 30, to = 90, by = 1))) %>% 
  data.frame() %>% 
  ggplot(aes(x = age,
             y = fit)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se),
              alpha = .2) +
  geom_line() +
  theme_bw() +
  labs(x = "Age, years",
       y = "Estimated Marginal Mean\nWeight, pounds")
```



