# D&H Ch6 - Statistical vs. Experimental Control: "PTSD"


Darlington & Hayes, Chapter 6's first example


```{r, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")
     
library(tidyverse) 
library(flextable)
library(apaSupp)
library(car)
library(rempsyc)
library(parameters)
library(performance)
library(interactions)
library(ggResidpanel)
```

```{r}
flextable::set_flextable_defaults(digits = 2)
```



## PURPOSE


```{block type='rmdlightbulb', echo=TRUE}
RESEARCH QUESTION:

Does the experimental therapy reduce PTSD symptom severity more than the traditional therapy?
```

### Data Description

Hypothetical study of the effect of a proposed therapy for Post-Traumatic Stress Disorder (PTSD) conducted on 14 military veterans who experienced combat.  Half were randomly assigned to experience 6 weeks of proposed therapy (experimental) and the other half received 6 weeks of traditional therapy.  Each was pre tested with respect to their PTSD symptoms and an identical post test assessment of their symptoms was administered upon the completion of the therapy.


* `pre_test` symptom severity prior to therapy
* `post_test` symptom severity after therapy
* `gain` symptom severity gain

* `therapy` which therapy was randomly assigned




Manually enter the data set provided on page 44 in Table 3.1



```{r}
df_ptsd <- tibble::tribble(~id, ~post_test, ~pre_test, ~therapy,
                            1,  2,   1,  1,  
                            2,  4,   3,  1,
                            3,  6,   7,  1,
                            4,  6,  10,  1,
                            5,  9,  13,  1,
                            6, 10,  17,  1,
                            7, 12,  19,  1,
                            8,  6,   1,  0,
                            9,  7,   5,  0,
                           10,  9,   7,  0,
                           11,  9,   9,  0,
                           12, 12,  13,  0,
                           13, 12,  16,  0,
                           14, 15,  19,  0) %>% 
  dplyr::mutate(therapy = factor(therapy,
                                 levels = c(0, 1),
                                 labels = c("Traditional", "Experimental"))) %>% 
  dplyr::mutate(gain = post_test - pre_test)
```


```{r}
df_ptsd %>% 
  dplyr::select("ID" = id,
                "Therapy"   = therapy,
                "Pre Test"  = pre_test,
                "Post Test" = post_test,
                "Gain"      = gain) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Dataset, wide Format") %>% 
  flextable::colformat_double(digits = 0)
```



### Longer Format of Repeated Measures

```{r}
df_ptsd_long <- df_ptsd %>% 
  dplyr::select(-gain) %>% 
  tidyr::pivot_longer(cols = ends_with("test"),
                      names_to = c("time", ".value"),
                      names_pattern = "(.*)_(.*)") %>% 
  dplyr::mutate(time = factor(time,
                              levels = c("pre", "post"))) %>% 
  dplyr::arrange(id, time)
```


```{r}
df_ptsd_long %>% 
  dplyr::select("ID" = id,
                "Therapy"    = therapy,
                "Time"       = time,
                "Test Score" = test) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Dataset, Long Format") %>% 
  flextable::colformat_double(digits = 0)
```



## EXPLORATORY DATA ANALYSIS

### Descriptive Statistics

#### Univariate


```{r}
df_ptsd %>% 
  dplyr::select("Therapy" = therapy) %>% 
  apaSupp::tab_freq(caption = "Summary of Categorical Measures")
```

```{r}
df_ptsd %>% 
  dplyr::select("Post Test" = post_test,
                "Pre Test"  = pre_test,
                "Gain" = gain) %>% 
  apaSupp::tab_desc(caption = "Summary of Measures")
```



#### Bivariate


```{r}
df_ptsd %>% 
  dplyr::select(therapy,
                "Post Test" = post_test,
                "Pre Test"  = pre_test,
                "Gain" = gain) %>% 
  apaSupp::table1_apa(split = therapy,
                      caption = "Descriptive Summary by Therapy Group")
```




```{block type='rmdimportant', echo=TRUE}
Pearson's correlation is only appropriate for TWO CONTINUOUS variables.  The exception is when 1 variable is continuous and the other has exactly 2 levels.  In this case, the binary variable needs to be converted to two numbers (numeric not factor) and the value is called the **Point-Biserial Correlation** ($r_{pb}$).
```


```{r}
df_ptsd %>% 
  dplyr::mutate(therapy = as.numeric(therapy == "Experimental")) %>% 
  dplyr::select("Post Test" = post_test,
                "Pre Test"  = pre_test,
                "Gain" = gain,
                "Therapy" = therapy) %>% 
  apaSupp::tab_cor(caption = "Correlation Between Pairs of Measures",
                   general_note = "For pairs of variables with therapy, r = Point-Biserial Correlation, otherwise ") %>% 
  flextable::hline(i = 3) %>% 
  flextable::bg(i = 1:3, bg = "yellow") %>% 
  flextable::bg(i = c(4), bg = "orange") 
```




### Visualizing Distributions

#### Univariate


```{r, fig.cap="Univariate Distibution of Measures"}
df_ptsd %>% 
  dplyr::mutate(therapy = as.numeric(therapy == "Experimental")) %>% 
  dplyr::select(id,
                "Post Symptom Severity" = post_test,
                "Pre Symptom Severity"  = pre_test,
                "Change in Symptom Severity" = gain,
                "Therapy" = therapy) %>%  
  tidyr::pivot_longer(cols = -id) %>% 
  ggplot(aes(value)) +
  geom_histogram(binwidth = 1,
                 color = "black",
                 alpha = .25) +
  theme_bw() +
  facet_wrap(~ name,
             scale = "free_x") +
  labs(x = NULL,
       y = "Count")
```


#### Bivariate


```{r, fig.cap="Distribution of Symptom Severity Prior to Therapy"}
apaSupp::spicy_histos(df = df_ptsd,
                      var = pre_test,
                      split = therapy,
                      lab = "Symptom Severity Prior to Therapy") +
  scale_x_continuous(breaks = seq(from = -24, to = 24, by = 2)) 
```



```{r, fig.cap="Distribution of Symptom Severity After Therapy"}
apaSupp::spicy_histos(df = df_ptsd,
                      var = post_test,
                      split = therapy,
                      lab = "Symptom Severity After Therapy") +
  scale_x_continuous(breaks = seq(from = -24, to = 24, by = 2))
```



```{r, fig.cap="Distribution of Change in Symptom Severity"}
apaSupp::spicy_histos(df = df_ptsd,
                      var = gain,
                      split = therapy,
                      lab = "Change in Symptom Severity") +
  scale_x_continuous(breaks = seq(from = -24, to = 24, by = 2)) 
```

```{r, fig.cap="Progression in Symptom Severity, Stratified by Therapy"}
df_ptsd_long %>% 
  ggplot(aes(x = time,
             y = test,
             group = id)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(x = NULL,
       y = "Symptom Severity Score") +
  facet_grid(~ therapy) +
  stat_summary(aes(group = 1),
           geom = "point",
           fun = "mean",
           shape = 13,
           size = 5,
           color = "red")+
  stat_summary(aes(group = 1),
           geom = "line",
           fun = "mean",
           color = "red",
           linewidth = 1.5)
```




#### Multivariate


```{r, fig.cap="Association Between Symptom Severity Prior to and After Therapy, Stratified by Therapy Type"}
df_ptsd %>% 
  ggplot(aes(x = pre_test,
             y = post_test)) +
  geom_point(aes(color = therapy,
                 shape = therapy),
             size = 3) + 
  theme_bw() +
  geom_smooth(aes(color = therapy,
                  fill = therapy,
                  linetype = therapy),
              method = "lm",
              formula = y ~x,
              alpha = .2) +
  labs(x = "Observed Symptom Severity Prior to Therapy",
       y = "Observed Symptom Severity\nAfter Therapy",
       color    = "Therapy Type:",
       fill     = "Therapy Type:",
       shape    = "Therapy Type:",
       linetype = "Therapy Type:") +
  theme(legend.position.inside = TRUE,
        legend.position = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_x_continuous(breaks = seq(from = 0, to = 24, by = 2)) +
  scale_y_continuous(breaks = seq(from = 0, to = 24, by = 2)) + 
  geom_abline(intercept = 0, slope = 1)
```



```{r, fig.cap="Association Between Test Gain in Score and Pre-Test, Stratified by Therapy"}
df_ptsd %>% 
  ggplot(aes(x = pre_test,
             y = gain)) +
  geom_point(aes(color = therapy,
                 shape = therapy),
             size = 3) + 
  theme_bw() +
  geom_smooth(aes(color = therapy,
                  fill = therapy,
                  linetype = therapy),
              method = "lm",
              formula = y ~x,
              alpha = .2) +
  labs(x = "Observed Symptom Severity Prior to Therapy",
       y = "Observed Change in Symptom Severity",
       color    = "Therapy Type:",
       fill     = "Therapy Type:",
       shape    = "Therapy Type:",
       linetype = "Therapy Type:") +
  theme(legend.position.inside = TRUE,
        legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_x_continuous(breaks = seq(from = 0, to = 24, by = 2))+
  scale_y_continuous(breaks = seq(from = -24, to = 24, by = 2)) +
  geom_hline(yintercept = 0)
```



## REGRESSION ANALYSIS

### DV = Post Score, controlling for Pre Score

```{r}
fit_ptsd_1 <- lm(post_test ~ pre_test + therapy,
                 data = df_ptsd)

summary(fit_ptsd_1)
```

```{r}
anova(fit_ptsd_1)
```




#### Equation

$$
\widehat{post} = 5.019 + 0.498(pre) -3.00(type=exp)
$$

#### Role of Therapy

Test the significance for therapy: *t*(11) = -8.33, *p* <.001, VERY significant

After controlling for pre-test score, the experimental therapy is associated with 3 point lower post-test score compared to the traditional therapy, *b* = -3.00, *SE* = 0.36, *p* <.001.

> Notes:
>
> Covariate `pre_test` has *b* = -0.50, *SE* = 0.03, *p* <.001. 
>
> The entire model: *SS*resid = 4.998, *R^2* = .970, *F*(2, 11) = 176.6, *p* < .001






### DV = Gain Score, ignoring Pre-Score

```{r}
fit_ptsd_2 <- lm(gain ~ therapy,
                 data = df_ptsd)

summary(fit_ptsd_2)
```


```{r}
anova(fit_ptsd_2)
```

#### Equation

$$
\widehat{gain} = 0 -3.00(EXP)
$$

#### Role of Therapy

Test the significance for therapy: *t*(12) = -1.67, *p* = .121, NOT significant

No evidence was found that the therapy type was associated with a differential change in symptom severity,*b* = -3.00, *SE* = 1.80, *p* =.121.

> Note:
>
> The entire model: *SS*resid = 136, *R^2* = .188, *F*(1, 11) = 2.78, *p* =.121



#### Equivalent Analysis: independent groups t-test

```{r}
t.test(gain ~ therapy,
       data = df_ptsd)
```




### DV = Gain Score, controlling for Pre-Score

```{r}
fit_ptsd_3 <- lm(gain ~ pre_test + therapy,
                 data = df_ptsd)

summary(fit_ptsd_3)
```


```{r}
anova(fit_ptsd_3)
```



#### Equation

$$
\widehat{gain} = 5.019 - 0.50(pre) -3.00(EXP)
$$

#### Role of Therapy

Test the significance for therapy type on symptom severity change: *t*(11) = -8.33, *p* <.001, VERY significant

After controlling for symptom severity prior to therapy, the experimental therapy was associated with 3 point lower change on symptom severity after therapy compared to traditional therapy, *b* = -3.00, *SE* = 0.36, *p* <.001.

> Notes:
>
> Covariate `pre_test` has *b* = -0.50, *SE* = 0.03, *p* <.001. 
>
> The entire model: *SS*resid = 4.998, *R^2* = .970, *F*(2, 11) = 178.8, *p* < .001





### Compare Models

```{r, warning=FALSE, message=FALSE}
performance::compare_performance(fit_ptsd_1, fit_ptsd_2, fit_ptsd_3,
                                 metrics = c("AIC", "BIC",
                                             "R2", "R2_adj",
                                             "RMSE", "SIGMA"))
```




## ALTERNATIVE ANALYSIS


### 2x2 mix ANOVA

```{r}
fit_aov <- afex::aov_4(test ~ therapy + (time|id),
                       data = df_ptsd_long)

summary(fit_aov)
```



