# D&H Ch7 - Regression for Prediction: "mtcars"


Darlington & Hayes, Chapter 7's  example


```{r, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("remotes")
# remotes::install_github("sarbearschwartz/apaSupp")
# remotes::install_github("ddsjoberg/gtsummary")
     
library(tidyverse) 
library(flextable)
library(apaSupp)
library(car)
library(rempsyc)
library(parameters)
library(performance)
library(interactions)
library(ggResidpanel)
```

```{r}
flextable::set_flextable_defaults(digits = 2)
```



## PURPOSE


```{block type='rmdlightbulb', echo=TRUE}
RESEARCH QUESTION:

What variables predict fuel efficiency?
```

### Data Description

Five candidate predictors of `mpg` (miles per gallon)


* `disp` displacement, cubic inches 
* `hp` gross hoursepower
* `drat` rear axle ratio
* `wt` weight, 1000s pounds
* `qsec` 1/4 mile time, seconds


```{r}
mtcars
```


## REGRESSION ANALYSIS


```{r}
fit_mtcars_1 <- lm(mpg ~ disp + hp + drat + wt + qsec, 
                   data = mtcars)

summary(fit_mtcars_1)
```

## AUTOMATED VARIABLE SELECTION

https://olsrr.rsquaredacademy.com/articles/variable_selection 


### All Possible Subsets

```{r}
fit_mtcars_1_all <- olsrr::ols_step_all_possible(fit_mtcars_1)

fit_mtcars_1_all
```


```{r}
plot(fit_mtcars_1_all)
```

### Best Subsets

```{r}
olsrr::ols_step_best_subset(fit_mtcars_1)
```

### Forward Stepwise, p-value

```{r}
olsrr::ols_step_forward_p(fit_mtcars_1)
```



### Backwards Stepwise, p-value

```{r}
olsrr::ols_step_backward_p(fit_mtcars_1)
```


### Forward Stepwise, R-squared

```{r}
olsrr::ols_step_forward_r2(fit_mtcars_1)
```







### Forward Stepwise, Adjusted R-squared

```{r}
olsrr::ols_step_forward_adj_r2(fit_mtcars_1)
```



### Forward Stepwise, AIC

```{r}
olsrr::ols_step_forward_aic(fit_mtcars_1)
```


### Forward Stepwise, Schwartz Bayesian Criterion

```{r}
olsrr::ols_step_forward_sbc(fit_mtcars_1)
```






