# Ex: Logistic - Maternal Risk (Hoffman)

Maternal Risk Factor for Low Birth Weight Delivery

Compiled: `r format(Sys.time(), '%B %d, %Y')`

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```


## PREPARATION

### Load Packages

```{r, message=FALSE, error=FALSE, warning=FALSE}

# library(remotes)
# remotes::install_github("sarbearschwartz/apaSupp")  # updated: 4/12/25
# remotes::install_github("ddsjoberg/gtsummary")


library(tidyverse)
library(haven)        
library(naniar)
library(apaSupp)
library(performance) 
library(interactions)
library(GGally)
```



### Load Data

More complex example demonstrating modeling decisions

Another set of data from a study investigating predictors of low birth weight

* `id` infant's unique identification number

Dependent variable (DV) or outcome    

* `low` Low birth weight (outcome) 
- 0 = birth weight >2500 g (normal)
- 1 = birth weight < 2500 g (low))  

* `bwt` actual infant birth weight in grams *(ignore for now)*


Independent variables (IV) or predictors

* `age` Age of mother, in years
* `lwt` Mother's weight at last menstrual period, in pounds
* `ftv` Number of physician visits in 1st trimester: 0 = None, 1 = One, ... 6 = six
* `race` Race: 1 = White, 2 = Black, 3 = Other
* `ptl` History of premature labor: 0 = None, 1 = One, 2 = two, 3 = three
* `ht` History of hypertension: 1 = Yes, 0 = No
* `smoke` Smoking status during pregnancy:1 = Yes, 0 = No
* `ui` Uterine irritability: 1 = Yes, 0 = No



The data is saved in a text file (`.txt`) without any labels.

```{r}
df_txt <- read.table("https://raw.githubusercontent.com/CEHS-research/data/master/Regression/lowbwt.txt", 
                     header = TRUE, 
                     sep = "", 
                     na.strings = "NA", 
                     dec = ".", 
                     strip.white = TRUE)

tibble::glimpse(df_txt)
```



### Wrangle Data


```{r}
df_mom <- df_txt %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(low = low %>% 
                  factor() %>% 
                  forcats::fct_recode("birth weight >2500 g (normal)" = "0",
                                      "birth weight < 2500 g (low)"   = "1")) %>% 
  dplyr::mutate(race = race %>% 
                  factor() %>% 
                  forcats::fct_recode("White" = "1",
                                      "Black" = "2",
                                      "Other" = "3")) %>% 
  dplyr::mutate(ptl_any = as.numeric(ptl > 0)) %>%         # collapse into 0 = none vs. 1 = at least one
  dplyr::mutate(ptl = factor(ptl)) %>%                     # declare the number of pre-term labors to be a factor: 0, 1, 2, 3
  dplyr::mutate_at(vars(smoke, ht, ui, ptl_any),           # declare all there variables to be factors with the same two levels
                   factor,
                   levels = 0:1,
                   labels = c("No", "Yes")) 
```


Display the structure of the 'clean' version of the dataset

```{r}
str(df_mom)
```

```{r}
tibble::glimpse(df_mom)
```




## EXPLORATORY DATA ANALYSIS

### Missing Data


```{r}
df_mom %>% 
  dplyr::select("Race" = race, 
                "Low Birth Weight Delivery" = low,
                "Age, years" = age, 
                "Weight, pounds" = lwt, 
                "1st Tri Dr Visits" = ftv, 
                "History of Premature Labor, any" = ptl_any, 
                "History of Premature Labor, number" = ptl, 
                "Smoking During pregnancy" = smoke, 
                "History of Hypertension" = ht, 
                "Uterince Irritability" = ui) %>% 
  naniar::miss_var_summary() %>%
  dplyr::select(Variable = variable,
                n = n_miss) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Missing Data by Variable")
```


### Summary


```{r}
df_mom %>% 
  dplyr::select("Low Birth Weight Delivery" = low,
                "Race" = race, 
                "Age, yrs" = age, 
                "Weight, lbs" = lwt, 
                "1st Tri Dr Visits" = ftv, 
                "Hx of Premature Labor, number" = ptl, 
                "Hx Premature Labor, any" = ptl_any, 
                "Hx of Hypertension" = ht, 
                "Smoking During pregnancy" = smoke, 
                "Uterince Irritability" = ui) %>% 
  apaSupp::tab_freq(caption = "Summary of Categorical Variables")
```



```{r}
df_mom %>% 
  dplyr::select("Low Birth Weight Delivery" = low,
                "Race" = race,  
                "Age, yrs" = age, 
                "Weight, lbs" = lwt, 
                "1st Tri Dr Visits" = ftv,
                "Hx Premature Labor, any" = ptl_any, 
                "Hx Premature Labor, number" = ptl, 
                "Hx Hypertension" = ht, 
                "Smoking During pregnancy" = smoke, 
                "Uterince Irritability" = ui) %>% 
  apaSupp::tab_desc(caption = "Summary of Continuous Variables")
```





## UNADJUSTED

> Unadjusted Models

### Fit Models

```{r}
fit_glm_race  <- glm(low ~ race,    family = binomial(link = "logit"), data = df_mom)
fit_glm_age   <- glm(low ~ age,     family = binomial(link = "logit"), data = df_mom)
fit_glm_lwt   <- glm(low ~ lwt,     family = binomial(link = "logit"), data = df_mom)
fit_glm_ftv   <- glm(low ~ ftv,     family = binomial(link = "logit"), data = df_mom)

fit_glm_ptl   <- glm(low ~ ptl_any, family = binomial(link = "logit"), data = df_mom)
fit_glm_ht    <- glm(low ~ ht,      family = binomial(link = "logit"), data = df_mom)
fit_glm_smoke <- glm(low ~ smoke,   family = binomial(link = "logit"), data = df_mom)
fit_glm_ui    <- glm(low ~ ui,      family = binomial(link = "logit"), data = df_mom)
```


### Parameter Tables


Note: the parameter estimates here are for the LOGIT scale, not the odds ration (OR) or even the probability.

```{r}
apaSupp::tab_glms(list(fit_glm_race, fit_glm_age, fit_glm_lwt, fit_glm_ftv),
                  narrow = TRUE,
                  fit = NA,
                  pr2 = "tjur")
```

```{r}
apaSupp::tab_glms(list(fit_glm_ptl, fit_glm_ht, fit_glm_smoke, fit_glm_ui),
                  narrow = TRUE,
                  fit = NA,
                  pr2 = "tjur")
```





## MAIN EFFECTS ONLY

Main-effects multiple logistic regression model

```{r}
fit_glm_mains <- glm(low ~ race + age + lwt + ftv + ptl_any + ht + smoke + ui,
                     family = binomial(link = "logit"), 
                     data = df_mom)
```


```{r}
apaSupp::tab_glm(fit_glm_mains,
<<<<<<< Updated upstream
                 show_single_row = c("smoke", "ptl_any", "ht", "ui"))
=======
                 var_labels = c(race = "Race",
                                age = "Age, yrs",
                                lwt = "Prior Weight, lbs",
                                ftv = "First Tri Visits",
                                ptl_any = "Hx Premature Labor",
                                ht = "Hx Hypertension",
                                smoke = "Smoking",
                                ui = "Uterine Irritability"),
                 show_single_row = c("ptl_any", "smoke", "ht", "ui"),
                 vif = FALSE) %>% 
  flextable::width(j = 1, width = 1.75)
>>>>>>> Stashed changes
```


## INTERACTIONS

Before removing non-significant main effects, test plausible interactions

Try the following interactions:

* Age and Weight  

* Age and Smoking

* Weight and Smoking

```{r}
fit_glm_mains_aw <- glm(low ~ race + age + lwt + ftv + ptl_any + ht + smoke + ui + age:lwt,
                        family = binomial(link = "logit"), 
                        data = df_mom)

fit_glm_mains_as <- glm(low ~ race + age + lwt + ftv + ptl_any + ht + smoke + ui + age:smoke,
                        family = binomial(link = "logit"), 
                        data = df_mom)

fit_glm_mains_ws <- glm(low ~ race + age + lwt + ftv + ptl_any + ht + smoke + ui + lwt:smoke,
                        family = binomial(link = "logit"), 
                        data = df_mom)
```



```{r}
apaSupp::tab_glms(list("Age-Weight"     = fit_glm_mains_aw,
                       "Age-Smoking"    = fit_glm_mains_as,
                       "Weight-Smoking" = fit_glm_mains_ws),
<<<<<<< Updated upstream
                  show_single_row = c("smoke", "ptl_any", "ht", "ui"),
                  narrow = TRUE)
=======
                  var_labels = c(race = "Race",
                                 age = "Age, yrs",
                                 lwt = "Prior Weight, lbs",
                                 ftv = "First Tri Visits",
                                 ptl_any = "Hx Premature Labor",
                                 smoke = "Smoking",
                                 ht = "Hypertension",
                                 ui = "Uterine Irritability"),
                  narrow = TRUE,
                  show_single_row = c("ptl_any", "smoke", "ht", "ui"))
>>>>>>> Stashed changes
```

```{r}
apaSupp::tab_glm_fits(list("Only Mains"     = fit_glm_mains,
                           "Age-Weight"     = fit_glm_mains_aw,
                           "Age-Smoking"    = fit_glm_mains_as,
                           "Weight-Smoking" = fit_glm_mains_ws))
```



```{r}
anova(fit_glm_mains, fit_glm_mains_aw, test = 'LRT')
```


```{r}
anova(fit_glm_mains, fit_glm_mains_as, test = 'LRT')
```


```{r}
anova(fit_glm_mains, fit_glm_mains_ws, test = 'LRT')
```








## PARSAMONY

No interactions are  significant
Remove non-significant main effects


Since the mother's age is theoretically a meaningful variable, it should probably be retained.

Remove "UI" since its not significant

```{r}
fit_glm_trim <- glm(low ~  race + age + lwt + ptl_any + ht + smoke ,
                    family = binomial(link = "logit"), 
                    data = df_mom)
```


```{r}
apaSupp::tab_glm(fit_glm_trim,
                 var_labels = c(race = "Race",
                                age = "Age, yrs",
                                lwt = "Prior Weight, lbs",
                                ptl_any = "Hx Premature Labor",
                                ht = "Hx Hypertension",
                                smoke = "Smoking"),
                 show_single_row = c("ptl_any", "smoke", "ht"),
                 lrt = FALSE,
                 vif = FALSE) %>% 
  flextable::width(j = 1, width = 1.75)
```




## CENTER & SCALE

Since the mother's age is theoretically a meaningful variable, it should probably be retained.

Revise so that age is interpreted in 5-year and pre-pregnancy weight in 20 lb increments and the intercept has meaning.



```{r}
fit_glm_final <- glm(low ~ race + I((age - 20)/5) + I((lwt - 125)/20) 
                     + ptl_any + ht + smoke ,
                     family = binomial(link = "logit"), 
                     data = df_mom)
```

```{r}
<<<<<<< Updated upstream
apaSupp::tab_glm(fit_glm_final,
                 var_labels = c("I((age - 20)/5)" = "Age, 5 yr",
                                "I((lwt - 125)/20)" = "Pre-Preg Weight, 20 lb",
                                race = "Race",
                                smoke = "Smoker",
                                ptl_any = "Prior Preterm Labor",
                                ht = "Hx Hypertension"),
                 p_note = "apa13",
                 show_single_row = c("smoke", "ptl_any", "ht"),
                 general_note = "Centering for age (20 yr) and weight (125 lbs)")
=======
apaSupp::tab_glm(fit_glm_trim,
                 var_labels = c(race = "Race",
                                age = "Age, yrs",
                                lwt = "Prior Weight, lbs",
                                ptl_any = "Hx Premature Labor",
                                ht = "Hx Hypertension",
                                smoke = "Smoking"),
                 show_single_row = c("ptl_any", "smoke", "ht"),
                 lrt = FALSE,
                 vif = FALSE) %>% 
  flextable::width(j = 1, width = 1.75)
>>>>>>> Stashed changes
```

## Probe & Plot


### focus: Compare Races with prior weight and hypertension

```{r}
<<<<<<< Updated upstream
interactions::interact_plot(model = fit_glm_final,
=======
interactions::interact_plot(model = fit_glm_trim,
>>>>>>> Stashed changes
                            pred = lwt,
                            modx = race,
                            legend.main = "Race",
                            mod2 = ht,
                            mod2.labels = c("History of Hypertension: No",
                                            "History of Hypertension: Yes"))  +
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(.5, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Mother's Weight Pre-Pregnancy, pounds",
       y = "Predicted Probability\nBaby has Low Birth Weight (< 2500 grams)",
       linetype = "Race") +
  scale_linetype_manual(values = c("solid", "longdash", "dotted"))
```



```{r, fig.width=8, fig.height=6}
<<<<<<< Updated upstream
effects::Effect(focal.predictors = c("age", "lwt", "race"),
                mod = fit_glm_final,
                xlevels = list(age = c(20, 30, 40),
                               lwt = seq(from = 80, to = 250, by = 5))) %>% 
  data.frame() %>% 
  dplyr::mutate(age_labels = glue::glue("Mother Age: {age}")) %>% 
=======
effects::Effect(mod = fit_glm_trim,
                focal.predictors = c("lwt", "race", "ptl_any", "ht", "smoke"),
                xlevels = list(lwt = seq(from = 80, to = 250, by = 5))) %>% 
  data.frame() %>%
  dplyr::filter(race != "Other") %>% 
  dplyr::mutate(risk = interaction(ptl_any, ht, smoke) %>% 
                  factor() %>% 
                  forcats::fct_recode("(0) None"               = "No.No.No",
                                      "(1) Preterm"            = "Yes.No.No" ,
                                      "(1) Ht"                 = "No.Yes.No" ,
                                      "(1) Smoking"            = "No.No.Yes" ,
                                      "(2) Preterm & Ht"       = "Yes.Yes.No" ,
                                      "(2) Ht & Smoking"       = "No.Yes.Yes" ,
                                      "(2) Preterm & Smoking"  = "Yes.No.Yes",
                                      "(3) All Three"          = "Yes.Yes.Yes") %>% 
                  forcats::fct_reorder(fit) %>% 
                  forcats::fct_rev()) %>%  
>>>>>>> Stashed changes
  ggplot(aes(x = lwt,
             y = fit)) +
  geom_line(aes(color = risk,
                linetype = risk,
                linewidth = risk)) +
  theme_bw() +
  facet_grid(~ race) +
  scale_linetype_manual(values = c("solid", 
                                   "longdash", "dotdash", "dashed", 
                                   "solid", "solid", "solid", 
                                   "solid")) +
  scale_linewidth_manual(values = c(1.5, 
                                    1.25, 1.25, 1.25, 
                                    .75, .75, .75, 
                                    .5)) +
  scale_color_manual(values = c("black", 
                                "red", "coral", "orange", 
                                "purple", "dodgerblue", "cyan3",  
                                "green4")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Mother's Weight Pre-Pregnancy, pounds",
       y = "Predicted Probability\nBaby has Low Birth Weight (< 2500 grams)",
       color    = "Risk Factors:",
       linetype = "Risk Factors:",
       linewidth= "Risk Factors:") +
  theme(#legend.position = "bottom",
        legend.key.width = unit(1.5, "cm")) 
```




#### Focus on: Mother's weight and smoking status during pregnancy, as well as history of any per-term labor and hypertension

<<<<<<< Updated upstream
```{r}
interactions::interact_plot(model = fit_glm_final,
                            pred = lwt,
                            modx = smoke,
                            mod2 = ptl_any)
```


```{r}
interactions::interact_plot(model = fit_glm_final,
                            pred = lwt,
                            modx = smoke,
                            mod2 = ht)
```
=======

> Illustates risk given the mother is 20 years old and white



>>>>>>> Stashed changes

```{r, fig.width=8, fig.height=6}
effects::Effect(mod = fit_glm_trim,
                focal.predictors = c("race", "lwt", "smoke", "ptl_any", "ht"),
                fixed.predictors = list(age = 20),
<<<<<<< Updated upstream
                mod = fit_glm_final,
=======
>>>>>>> Stashed changes
                xlevels = list(lwt = seq(from = 80, to = 250, by = 5))) %>% 
  data.frame() %>% 
  dplyr::filter(race == "White") %>% 
  dplyr::mutate(smoke = forcats::fct_rev(smoke)) %>% 
  dplyr::mutate(ptl_any_labels = glue::glue("History of Preterm Labor: {ptl_any}")) %>% 
  dplyr::mutate(ht_labels = glue::glue("History of Hypertension: {ht}") %>% forcats::fct_rev()) %>% 
  ggplot(aes(x = lwt,
             y = fit)) +
  geom_line(aes(color = smoke,
                linetype = smoke),
            linewidth = 1) +
  theme_bw() +
  facet_grid(ht_labels ~ ptl_any_labels) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Mother's Weight Pre-Pregnancy, pounds",
       y = "Predicted Probability\nBaby has Low Birth Weight (< 2500 grams)",
       color    = "Mother Smoked",
       linetype = "Mother Smoked") +
  theme(legend.position = "inside",
        legend.position.inside = c(1, .5),
        legend.justification = c(1.1, 1.15),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm")) +
  scale_linetype_manual(values = c("longdash", "solid")) +
  scale_color_manual(values = c( "coral2", "dodger blue"))
```

