# Example: Obesity and Blood Pressure


## Packages

```{r, comment=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(furniture)
library(texreg)
library(stargazer)
library(psych)
library(car)
library(gpairs)
library(GGally)
library(corrplot)
library(ggthemes)
library(RColorBrewer)
library(effects)
library(ISwR)  # Introduction to Statistics with R (datasets)
```



## The Data


Load dataset that is included in the `ISwR` package

```{r}
data(bp.obese)
```

View documentation for the dataset (required for all data included in packages) by typeing `?bp.obese` in the console

Have a quick *look* at what we have in the data:


```{r}
glimpse(bp.obese)
```

### Declare and Label Factors

```{r}
bp.obese <- bp.obese %>% 
  dplyr::mutate(sex = factor(sex,
                             labels = c("Male", "Female"))) 

glimpse(bp.obese)
```

## Explore the Raw Data

### Tabular Descriptions: Summary Statistics

```{r, results='asis'}
stargazer::stargazer(bp.obese,
                     type = "html")  
```



```{r, results='asis'}
furniture::table1(bp.obese,
                  output = "html")  
```



```{r, results='asis'}
bp.obese %>% 
  furniture::table1(obese, bp,
                    splitby = ~ sex,
                    test = TRUE,
                    output = "html")  
```




### Visualization

`base` graphics:

```{r}
pairs(bp.obese)
```

`psych` package:

```{r}
psych::pairs.panels(bp.obese)
```


`gpairs` package:

```{r}
gpairs::gpairs(bp.obese)
```

Change up the defaults:

```{r}
gpairs::gpairs(bp.obese, 
               upper.pars   = list(scatter = "stats"),
               scatter.pars = list(col = as.numeric(bp.obese$sex)))
```


`GGally` package (wrapper for `ggplot2`):

```{r}
GGally::ggpairs(bp.obese)
```


Use the options to make it even better:

```{r}
# fancy options (I can't figure out how to change the color palette)
GGally::ggpairs(bp.obese,
                mapping = aes(fill      = sex,
                              col       = sex,
                              alpha     = 0.1),
                upper = list(continuous = "smooth",
                             combo      = "facethist",
                             discrete   = "ratio"),
                lower = list(continuous = "cor",
                             combo      = "box",
                             discrete   = "facetbar"),
                title = "Very Useful for Exploring Data") 
```




`ggplot2`: specify plot type

```{r}
ggplot(data = bp.obese, 
       mapping = aes(x = sex, y = bp)) + 
  geom_boxplot() 
```

`ggplots` can be assigned (stored to an object and refered to by name)

```{r}
fig_bp_by_sex <- bp.obese %>% 
  ggplot() +
  aes(x    = sex, 
      y    = bp,
      fill = sex) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_manual(values = c("mediumblue", "maroon3")) +
  labs(x = "Gender",
       y = "Blood Pressure (mmHg)") +
  guides(fill = FALSE) +
  theme_bw()
```

Call the plot by just the name it was assigned

```{r}
fig_bp_by_sex
```

Add onto a plot...change to a theme from the `ggplot2` package
```{r}
fig_bp_by_sex + theme_gray()        # 'gray' is the default
fig_bp_by_sex + theme_bw()
fig_bp_by_sex + theme_classic()
fig_bp_by_sex + theme_dark()
fig_bp_by_sex + theme_minimal()
fig_bp_by_sex + theme_linedraw()
```

add onto a plot...change to a theme from the `ggthemes` package

```{r}
fig_bp_by_sex + theme_tufte()             # Ed Tufte - author, minimalist
fig_bp_by_sex + theme_stata()             # Stata is a stats software
fig_bp_by_sex + theme_economist()         # 'The Economist' magazine
fig_bp_by_sex + theme_economist_white()
fig_bp_by_sex + theme_wsj()               # 'The Wall Street Journal
fig_bp_by_sex + theme_calc()              # based on Google Docs
fig_bp_by_sex + theme_hc()   
```

Visual inspection for an interaction (is gender a moderator?)

```{r}
fig_bp_by_obese <- bp.obese %>% 
  ggplot(aes(x     = obese,
             y     = bp,
             color = sex)) +
  geom_point(size  = 3)

fig_bp_by_obese
```

Can save as: pdf, jpeg, tiff, png, bmp, ... (see help) 

```{r, eval=FALSE}
ggsave("fig_bp_by_obese.png",     # name of file w/extension 
       fig_bp_by_obese,           # name of the ggplot object
       width  = 5,                # size, width
       height = 5,                # size, height
       units  = "in")             # size, units for width & height
```

add a smoother...the regression line with a conf bands for the mean

```{r}
fig_bp_by_obese + geom_smooth(method = "lm")
```


create seperate panels
```{r}
fig_bp_by_obese + facet_grid(. ~ sex) 
```


```{r}
fig_bp_by_obese + 
  facet_grid(. ~ sex) + 
  geom_smooth(method = "lm", level = .95)
```

get fancy

```{r}
fig_bp_by_obese + 
  facet_grid(. ~ sex) + 
  geom_smooth(method = "lm",
              se     = FALSE,
              size   = 0.9) +
  theme_bw() +
  guides(color = FALSE) +
  labs(title = "Does Gender Moderated the Obese-Blood Pressure Assocation?") +
  scale_color_manual(values = c("mediumblue", "maroon3")) 
```

get even fancier

```{r}
fig_bp_by_obese +
  geom_smooth(aes(fill = sex),
              alpha  = 0.2,
              method = "lm") +
  scale_color_manual(values = c("mediumblue", "maroon3"),
                     breaks = c("male",       "female"),
                     labels = c("Men",        "Women")) +
  scale_fill_manual(values  = c("mediumblue", "maroon3"),
                    breaks  = c("male",       "female"),
                    labels  = c("Men",        "Women")) +
  labs(title = "Does Gender Moderate the Association Between Obesity and Blood Pressure?",
       x     = "Ratio: Actual Weight vs. Ideal Weight (NYM Life Tables)",
       y     = "Systolic Blood Pressure (mmHg)") + 
  theme_bw() +
  scale_x_continuous(breaks  = seq(from = 0,  to = 3,   by = 0.25 )) +
  scale_y_continuous(breaks  = seq(from = 75, to = 300, by = 25)) +
  theme(legend.title         = element_blank(),
        legend.key           = element_rect(fill = "white"),
        legend.background    = element_rect(color = "black"),
        legend.justification = c(1, 0), 
        legend.position      = c(1, 0))
```




### Codrrelation: pairwise, un-adjusted 


calculate the correlation - nested format
```{r}
cor(as.numeric(bp.obese$sex), bp.obese$bp)
```

calculate the correlation - with the pipe

```{r}
bp.obese %>% 
  dplyr::mutate(sex = as.numeric(sex)) %>%  # cor needs only numeric
  cor() 
```


```{r}
bp.obese %>% 
  dplyr::mutate(sex = as.numeric(sex)) %>%  # cor needs only numeric
  dplyr::select(sex, bp) %>%                # limit to only 2 vars
  cor()                                # optional () on end if empty
```


```{r}
bp.obese %>% 
  dplyr::mutate(sex = as.numeric(sex)) %$%  # allows you to call vars by name
  cor(sex, bp)                            # just 'var' without the 'data$'
```



addign the correlation matrix to a named object

```{r}
cor_bp <- bp.obese %>% 
  dplyr::mutate(sex = as.numeric(sex)) %>% 
  cor() 

cor_bp
```


```{r}
cor_bp %>% round(3)
```

### Correlograms - visually display correlation matrix



all defaults
```{r}
corrplot::corrplot(cor_bp)
```

                   

change the visualization glyphs   
```{r}
corrplot::corrplot(cor_bp, method = "ellipse") 
```


```{r}
corrplot::corrplot(cor_bp, method = "pie")   
```


```{r}    
corrplot::corrplot(cor_bp, method = "color") 
```


```{r}
corrplot::corrplot(cor_bp, method = "number") 
```

full matrix, or only the upper or lower half shown
```{r}
corrplot::corrplot(cor_bp, type = "upper") 
```

```{r}
corrplot::corrplot(cor_bp, type = "lower") 
```


Brewer color palettes

```{r}
display.brewer.all()
```



change the colors used on the glyphs

```{r}
corrplot::corrplot(cor_bp, col = brewer.pal(n = 8,  name = "RdBu")) 
```

```{r}
corrplot::corrplot(cor_bp, col = brewer.pal(n = 6,  name = "RdYlBu")) 
```

```{r}
corrplot::corrplot(cor_bp, col = brewer.pal(n = 10, name = "PuOr")) 
```

get a bit fancier

```{r}
corrplot::corrplot.mixed(cor_bp)
```

```{r}
corrplot::corrplot.mixed(cor_bp,
                       lower  = "ellipse",
                       upper  = "number",
                       tl.col = "black")
```



## Multiple Linear Regression (MLR)


assign the model to an object you can call by name
```{r}
fit_bp_sex <- lm(bp ~ sex, data = bp.obese) # fit the model
```


###Call the model object several ways

what does the "lm" function save for you?

```{r}
attributes(fit_bp_sex)
```

what output does calling the name print?

```{r}
fit_bp_sex 
```

this call gives you MORE output

```{r}
summary(fit_bp_sex) 
```

you can access just the betas 

```{r}
coef(fit_bp_sex) 
```

the CI's for the betas

```{r}
confint(fit_bp_sex)
```

you can call just the ANOVA table

```{r}
anova(fit_bp_sex) 
```

### Model fit indicies

-2LL to compare nested models
```{r}
logLik(fit_bp_sex) 
```

```{r}
AIC(fit_bp_sex)
```

```{r}
BIC(fit_bp_sex)
```


### Tablular Results

```{r, results='asis'}
stargazer::stargazer(fit_bp_sex,
                     type = "html")
```


```{r, results='asis'}
stargazer(fit_bp_sex, 
          type = "html", 
          single.row = TRUE, 
          ci = TRUE)
```



### Fit Nested Models

```{r}
fit_bp_obe    <- lm(bp ~ obese,                   data = bp.obese)
fit_bp_obesex <- lm(bp ~ obese + sex,             data = bp.obese)
fit_bp_inter  <- lm(bp ~ obese + sex + obese*sex, data = bp.obese)
```

```{r, results='asis'}
stargazer::stargazer(fit_bp_sex, fit_bp_obe, fit_bp_obesex, fit_bp_inter,
                     type = "html",
                     title = "Model Comparison",
                     intercept.bottom = FALSE,
                     keep.stat = c("rsq", "adj.rsq", "aic"))
```


### Test Assumptions w/ residual diagnostics


The residual plots are hard to use when the single IV is binary

using `base` R 
```{r}
plot(fit_bp_sex)  
```

```{r}
plot(fit_bp_sex, which = 1)  # Residuals vs Fitted
plot(fit_bp_sex, which = 2)  # Normal Q-Q
plot(fit_bp_sex, which = 3)  # Scale-Location
plot(fit_bp_sex, which = 4)  # Cook's distance
plot(fit_bp_sex, which = 5)  # Residuals vs Leverage
plot(fit_bp_sex, which = 6)  # Cook's dist vs Leverage
```


```{r}
plot(fit_bp_sex, which = 4, id.n = 10)  # Change the number labeled
```

using the `car` package
```{r}
car::residualPlots(fit_bp_sex)    
```

you can adjust any part of a ggplot
```{r}
bp.obese %>% 
  dplyr::mutate(e_bp = resid(fit_bp_sex)) %>%  # add the resid to the dataset
  ggplot(aes(x     = sex,               # x-axis variable name
             y     = e_bp,              # y-axis variable name
             color = sex,               # color is the outline
             fill  = sex)) +            # fill is the inside
  geom_hline(yintercept = 0,               # set at a meaningful value
             size       = 1,               # adjust line thickness
             linetype   = "dashed",        # set type of line
             color      = "purple") +      # color of line
  geom_boxplot(alpha = 0.5) +                # level of transparency
  theme_bw() +                               # my favorite theme
  labs(title = "Check Assumptions",            # main title's text
       x = "Gender",                           # x-axis text label
       y = "Blood Pressure, Residual (bpm)") + # y-axis text label
  scale_y_continuous(breaks = seq(from = -40,    # declare a sequence of
                                  to   =  80,    # values to make the 
                                  by   =  20)) + # tick marks at
  guides(color = FALSE, fill = FALSE)               # no legends included
```

If you don't specify a name of a ggplot object, it will save the last one created (see 'Plots' tab)
 
```{r, eval=FALSE} 
ggsave("fig_resid-bp_by_obese.png",  # name of file w/extension 
       width  = 5,                   # size, width
       height = 5,                   # size, height
       units  = "in")                # size, units for width & height
```

```{r}
bp.obese %>% 
  dplyr::mutate(e_bp = resid(fit_bp_sex)) %>%  # add the resid to the dataset
  ggplot(aes(x     = e_bp,              # y-axis variable name
             color = sex,               # color is the outline
             fill  = sex)) +            # fill is the inside
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0,               # set at a meaningful value
             size       = 1,               # adjust line thickness
             linetype   = "dashed",        # set type of line
             color      = "purple") +      # color of line
  theme_bw() +                               # my favorite theme
  labs(title = "Check Assumptions",            # main title's text
       x = "Blood Pressure, Residual (bpm)") + # y-axis text label
  scale_x_continuous(breaks = seq(from = -40,    # declare a sequence of
                                  to   =  80,    # values to make the 
                                  by   =  20))  # tick marks at
```








