# Interactions Example

```{r}
library(tidyverse)
library(emmeans)
library(furniture)
library(texreg)
```


Taken from: https://stats.idre.ucla.edu/r/seminars/interactions-r/

## Motivation

Suppose you are doing a simple study on weight loss and notice that people who spend more time exercising lose more weight. Upon further analysis you notice that those who spend the same amount of time exercising lose more weight if they are more effortful. The more effort people put into their workouts, the less time they need to spend exercising. This is popular in workouts like high intensity interval training (HIIT).

You know that hours spent exercising improves weight loss, but how does it interact with effort? Here are three questions you can ask based on hypothetical scenarios.

* I’m just starting out and don’t want to put in too much effort. How many hours per week of exercise do I need to put in to lose 5 pounds?

* I’m moderately fit and can put in an average level of effort into my workout. For every one hour increase per week in exercise, how much additional weight loss do I expect?

* I’m a crossfit athlete and can perform with the utmost intensity. How much more weight loss would I expect for every one hour increase in exercise compared to the average amount of effort most people put in?

Additionally, we can *visualize the interaction* to help us understand these relationships.

## The Weight Loss Study

This is a hypothetical study of weight loss for 900 participants in a year-long study of 3 different exercise programs, a jogging program, a swimming program, and a reading program which serves as a control activity. 

**Variables** 

* `loss`: weight loss (continuous), positive = weight loss, negative scores = weight gain
* `hours`: hours spent exercising *(continuous)*
* `effort`: effort during exercise *(continuous)*
    + 0 = minimal physical effort and 
    + 50 = maximum effort
* `prog`: exercise program *(categorical)*
    + jogging=1
    + swimming=2
    + reading=3
* `gender`: participant gender *(binary)*
    + male=1
    + female=2


**Definitions**

> What exactly do I mean by decomposing, probing, and plotting an interaction?

* **decompose**: to break down the interaction into its lower order components (i.e., predicted means or simple slopes)
* **probe**: to use hypothesis testing to assess the statistical significance of simple slopes and simple slope differences (i.e., interactions)
* **plot**: to visually display the interaction in the form of simple slopes such as values of the dependent variable are on the y-axis, values of the predictor is on the x-axis, and the moderator separates the lines or bar graphs


Let’s define the **essential elements** of the interaction in a regression:

* DV: **dependent variable** (Y), the outcome of your study (e.g., weight loss)
* IV: **independent variable** (X), the predictor of your outcome (e.g., time exercising)
* MV: **moderating variable** (W) or moderator, a predictor that changes the relationship of the IV on the DV (e.g, effort)
* **coefficient**: estimate of the direction and magnitude of the relationship between an IV and DV
* **continuous variable**: a variable that can be measured on a continuous scale, e.g., weight, height
* **categorical or binary variable**: a variable that takes on discrete values, binary variables take on exactly two values, categorical variables can take on 3 or more values (e.g., gender, ethnicity)
* **main effects or slopes**: effects  or slopes for models that do not involve interaction terms
* **simple slope**: when a continuous IV interacts with an MV, its slope at a particular level of an MV
* **simple effect**: when a categorical IV interacts with an MV, its effect at a particular level of an MV


### Import Data

```{r}
data_raw <- read.csv("https://stats.idre.ucla.edu/wp-content/uploads/2019/03/exercise.csv")
```



```{r}
tibble::glimpse(data_raw)
```

```{r}
data_raw %>% 
  summary()
```
### Wrangle the data 

```{r}
data_clean <- data_raw %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(gender = factor(gender) %>% 
                  forcats::fct_recode("Male"   = "1",
                                      "Female" = "2")) %>% 
  dplyr::mutate(prog = factor(prog) %>% 
                  forcats::fct_recode("Jog"  = "1",
                                      "Swim" = "2",
                                      "Read" = "3"))
```


```{r}
tibble::glimpse(data_clean)
```

```{r}
summary(data_clean)
```

## Exploratory Data Analysis

### Summary Statistics

```{r}
data_clean %>% 
  dplyr::group_by(prog) %>% 
  furniture::table1("Gender" = gender,
                    "Weight Loss, pounds" = loss, 
                    "Hours Spent in Program" = hours, 
                    "Effort in Program" = effort,
                    digits = 2,
                    output = "markdown",
                    total = TRUE)
```

```{r}
data_clean %>% 
  dplyr::select_if(is.numeric) %>% 
  furniture::tableC()
```

```{r}
data_clean %>% 
  cor.test(~ loss + hours,
           data = .)
```


```{r}
data_clean %>% 
  cor.test(~ loss + effort,
           data = .)
```

```{r}
data_clean %>% 
  cor.test(~ hours + effort, 
           data = .)
```

### Visualization

```{r}
data_clean %>% 
  ggplot(aes(x = hours,
             y = loss)) +
  geom_point() +
  theme_bw()
```
```{r}
data_clean %>% 
  ggplot(aes(x = hours,
             y = loss)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

```{r}
data_clean %>% 
  ggplot(aes(x = hours,
             y = loss)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  facet_grid(~ gender)
```
```{r}
data_clean %>% 
  ggplot(aes(x = hours,
             y = loss)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  facet_grid(~ prog)
```
```{r}
data_clean %>% 
  ggplot(aes(x = hours,
             y = loss)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  facet_grid(gender ~ prog)
```



## Simple Linear Regression

### Fit

**Question:** Does time spent on a program effect weight loss?

```{r}
fit_lm_loss_hr <- lm(loss ~ hours,
                     data = data_clean)
```


```{r}
summary(fit_lm_loss_hr)
```

**Answer:** Yes.  There is evidence that an additional hour in a program is associated with slightly more weight lost, b = 2.47, t(1) = 2.61, p = .009, R^2 < .01.


### Predict

```{r}
fit_lm_loss_hr %>% 
  emmeans::emmeans(~ hours)
```

**Interpretation:** A participant that spends 2 hours on their program looses 10 pounds on average, SE = 0.47, 95% CI [9.1, 10.9].


```{r}
fit_lm_loss_hr %>% 
  emmeans::emmeans(~ hours,
                   at = list(hours = 1:4))
```

### Overall Slope

```{r}
fit_lm_loss_hr %>% 
  emmeans::emtrends(~ 1,
                    var = "hours")
```


**Interpretation:** Overall, on average the effect of time is an additional 2.47 pounds lost per hour spent on a program.  


### Plot

```{r}
fit_lm_loss_hr %>% 
  emmeans::emmip(~ hours,
                 at = list(hours = seq(from = .25, to = 4, by = .25)),
                 CIs = TRUE) +
  labs(caption = "Note: Confidence bands display 97% CI for the mean")
```


##  Continuous by Continuous

### Fit


The same model:
* `loss ~ hours*effort`
* `loss ~ hours + effore + hours:effort`


**Quesiton:** Is there an interaction between time spent and effort in a weight loss program?

```{r}
fit_lm_loss_hr_eff <- lm(loss ~ hours*effort,
                         data = data_clean)
```


```{r}
summary(fit_lm_loss_hr_eff)
```

**Answer:** Effect does moderate time spent in a program for weight loss, b = 0.39, t(1) = 2.10, p = .036.  The combined effect of time and effort account for nearly 8% of the variation in pounds lost.

### Table

Sometimes its nice to have a table of the parameter estimates of competing models.

```{r}
texreg::screenreg(list(fit_lm_loss_hr, 
                       fit_lm_loss_hr_eff),
                  single.row = TRUE)
```


### Predict

By default, we get a prediction for the mean hours (2) and the mean effort (29.7).

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emmeans(~ hours*effort)
```

We can customize which hours and effor for which we want to get predictions for.

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emmeans(~ hours*effort,
                   at = list(hours = 1:4,
                             effort = c(20, 30, 40)))
```

By adding the word `pairwise` we also get pairwise t-tests! Tukey's HSD adjustment for multiple compairisons is done by default.

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emmeans(pairwise ~ hours*effort,
                   at = list(hours = 2:3,
                             effort = c(25, 35)))
```

By changing the astrics (`*`) to a vertical bar (`|`) we get compairisons between different hours WITHIN each effort level.

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emmeans(pairwise ~ hours|effort,
                   at = list(hours = c(1, 4),
                             effort = c(20, 30, 40)))
```


### Simple Slope

By default, we only get the overall slope for time spent for the mean effort (29.7).

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emtrends(~ effort,
                    var = "hours")
```


Of course, you can customize the effort levels for which you want the slope of time spent.  These are called "simple slopes".

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emtrends(~ effort,
                    var = "hours",
                    at = list(effort = c(20, 30, 40)))
```

Again, adding the word `pairwise` will cause pairwise t-Tests to be ran.  Here I've asked NOT to do Tukey's HSD adjustment to show that is can be done.

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emtrends(pairwise ~ effort,
                    var = "hours",
                    at = list(effort = c(20, 30, 40)),
                    adjust = "none")
```


### Plot

The `emmeans` package can make interaction plots with the `emmip()` function, but it only can create confidene intervals, not confidence bands.

```{r}
fit_lm_loss_hr_eff %>% 
  emmeans::emmip(effort ~ hours,
                 at = list(hours = seq(from = .25, to = 4, by = .25),
                           effort = c(20, 30, 40)),
                 CIs = TRUE)
```

The `interactions` package can create bands with the `interact_plot()` function.

```{r}
fit_lm_loss_hr_eff %>% 
  interactions::interact_plot(pred = hours,
                              modx = effort,
                              interval = TRUE,
                              int.type = "confidence")
```


Since these packages are built on `ggplot2` you can customise them further to make them 'better' for publication.

```{r}
fit_lm_loss_hr_eff %>% 
  interactions::interact_plot(pred = hours,
                              modx = effort,
                              interval = TRUE,
                              int.type = "confidence") +
  labs(x = "Hours Spent Exercising",
       y = "Weight Loss, pounds",
       color = "Effort",
       fill = "Effort",
       linetype = "Effort",
       alpha = "Effort") +
  theme_bw() +
  theme(legend.position = c(0, 1),
        legend.justification = c(-0.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```

**Interpretation:** When little time is spent on the program, participants lost just over 5 pounds, irrespect of effort.  More time spent on the programs only translated to additional weight lost if their effort was high.


## Continuous by Categorical

### Fit

**Question:** Did weight loss depend on gender?

```{r}
fit_lm_loss_gen <- lm(loss ~ gender,
                      data = data_clean)
```

```{r}
summary(fit_lm_loss_gen)
```
**Answer:**The main effect of gender is not significant.


**Question:** Does gender moderate the effect of spending addition time on the program? *(ignoring the role of effort and program type for the time being)*

```{r}
fit_lm_loss_hrs_gen <- lm(loss ~ hours*gender,
                          data = data_clean)
```


```{r}
summary(fit_lm_loss_hrs_gen)
```

**Answer:** No, gender does not interact with time spent.  

**Question:** Is the effect of time spend moderated by type of program?

```{r}
fit_lm_loss_hrs_prog <- lm(loss ~ hours*prog,
                          data = data_clean)
```


```{r}
summary(fit_lm_loss_hrs_prog)
```
```{r}
anova(fit_lm_loss_hrs_prog)
```


**Answer:** Yes!  The type of program does moderate the effect of time spent on weight loss, F(2, 894) = 62.91, p < .001.


### Table

Sometimes its nice to have a table of the parameter estimates of competing models.

```{r}
texreg::screenreg(list(fit_lm_loss_hr, 
                       fit_lm_loss_gen,
                       fit_lm_loss_hrs_gen,
                       fit_lm_loss_hrs_prog),
                  custom.model.names = c("Hrs",
                                         "Gender",
                                         "Hrs + Gender",
                                         "Hrs + Program"))
```

### Predict

```{r}
fit_lm_loss_hrs_prog %>% 
  emmeans::emmeans(~hours)
```

```{r}
fit_lm_loss_hrs_prog %>% 
  emmeans::emmeans(~ prog)
```

```{r}
fit_lm_loss_hrs_prog %>% 
  emmeans::emmeans(~ hours|prog)
```

```{r}
fit_lm_loss_hrs_prog %>% 
  emmeans::emmeans(~ hours|prog,
                   at = list(hours = 1:4))
```



### Simple Slopes

```{r}
fit_lm_loss_hrs_prog %>% 
  emmeans::emtrends(pairwise ~ prog,
                    var = "hours")
```

**Interpretation:** Participants in the jogging program (M = 7.45 lb/hr, SE = 0.81) and the swimming program (M = 6.87 lb/hr, SE = 0.78), loose about the same amount of weight for each additional hour spent exercising, p = .863, conversly, each additional reading is associated with more weight gained (M = -2.96 lb/hr, SE = 0.71). 

### Plot

```{r}
fit_lm_loss_hrs_prog %>% 
  interactions::interact_plot(pred = hours,
                              modx = prog,
                              interval = TRUE,
                              int.type = "confidence")
```


## Categorical by Categorical

### Fit

**Question:** Do both genders have the same success in all the programs?

```{r}

fit_lm_loss_prog <- lm(loss ~ prog,
                       data = data_clean)

fit_lm_loss_prog_gen <- lm(loss ~ prog*gender,
                           data = data_clean)
```


```{r}
summary(fit_lm_loss_prog_gen)
```

```{r}
anova(fit_lm_loss_prog_gen)
```
**Answer:** At least one of the programs results in more weight lost for one of the genders. 


### Table

```{r}
texreg::screenreg(list(fit_lm_loss_gen,
                       fit_lm_loss_prog,
                       fit_lm_loss_prog_gen),
                  custom.model.names = c("Gender",
                                         "Program",
                                         "Interacting"))
```


### Predict

**Question:** Which gender or program looses the most and least weight?

```{r}
fit_lm_loss_prog_gen %>% 
  emmeans::emmeans(pairwise ~ prog | gender)
```

**Answer:** For each gender, the swimming program resultsed in the most weight lost, followed by the jogging program, where as the the reading program paticipants gained weight.


```{r}
fit_lm_loss_prog_gen %>% 
  emmeans::emmeans(pairwise ~ gender | prog)
```

**Answer:** In the swimming program females lost more weight than males, but the reverse trend was observed in the jogging group.  In the reading group there was no gender gap in weight lost.



### Simple Slopes

> Not appropriate for categorical-only interactions


### Plot

```{r}
fit_lm_loss_prog_gen %>% 
  emmeans::emmip(gender ~ prog,
                 CIs = TRUE)
```
```{r}
fit_lm_loss_prog_gen %>% 
  emmeans::emmip(prog ~ gender,
                 CIs = TRUE)
```


## Three-way Interaction

### Fit



```{r}
fit_lm_loss_3way_a <- lm(loss ~ prog*gender*effort,
                           data = data_clean)
```


```{r}
summary(fit_lm_loss_3way_a)
```
```{r}
anova(fit_lm_loss_3way_a)
```

```{r}
fit_lm_loss_3way_b <- lm(loss ~ prog*gender*hours,
                           data = data_clean)
```


```{r}
summary(fit_lm_loss_3way_b)
```



```{r}
anova(fit_lm_loss_3way_b)
```

### Table

We can ask the table to show confidence intervals (95% by default) instead of standard errors.

```{r}
texreg::screenreg(list(fit_lm_loss_3way_b),
                  single.row = TRUE,
                  ci.force = TRUE)
```



### Plot

```{r}
fit_lm_loss_3way_b %>% 
  interactions::interact_plot(pred = hours,            # x-axis
                              modx = prog,             # separate lines
                              mod2 = gender,           # separate panels
                              interval = TRUE,         # add bands
                              int.type = "confidence")
```


```{r}
fit_lm_loss_3way_b %>% 
  interactions::interact_plot(pred = hours,            # x-axis
                              modx = gender,           # separate lines
                              mod2 = prog,             # separate panels
                              interval = TRUE,         # add bands
                              int.type = "confidence")
```
