# Interactions Example

```{r, comment=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
library(emmeans)
library(furniture)
library(stargazer)
library(psych)
library(texreg)
library(interactions)
library(rgl)
```


```{r, echo=FALSE}
knitr::knit_hooks$set(webgl = hook_webgl)
```



Regression Analysis and Linear Models: concepts, applicaitons, and implementation
- By Richard B. Darlington & Andrew F. Hayes

Chapter 3: Partial Relationships and the Multiple Linear Regressio Model


## The Data

Data from Table 3.1

```{r}
df <- data.frame(id = 1:10,
                 exercise = c( 0,  0,  0,  2,  2,  2,  2,  4,  4,  4),
                 food     = c( 2,  4,  6,  2,  4,  6,  8,  4,  6,  8),
                 metabo   = c(15, 14, 19, 15, 21, 23, 21, 22, 24, 26),
                 loss     = c( 6,  2,  4,  8,  9,  8,  5, 11, 13,  9))

df
```

## Visualize

Figure 3.1

An example with a positive simple association and negative partial association.


```{r}
df %>% 
  dplyr::mutate(exercise = factor(exercise)) %>% 
  ggplot(aes(x = food,
             y = loss,
             color = exercise,
             shape = exercise)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(x = "Food Intake, 100s of calories above recommended",
       y = "Weight Loss, 100s of grams",
       color = "Exercise Frequency, average weekly hours: ",
       shape = "Exercise Frequency, average weekly hours: ") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  scale_shape_manual(values = c(15, 19, 17))  +
  geom_smooth(aes(group = 1),
              method = "lm", 
              se = FALSE,
              color = "black") +
  geom_smooth(aes(group = exercise),
                  method = "lm", 
              se = FALSE,
              linetype = "dashed")
```


## Regression

GEneric Form

$$
Y_i = b_0 + b_1X_{1i} + b_2X_{2i}+ e_i
$$


$$
\hat{Y} = b_0 + b_1X_1 + b_2X_2
$$

```{r}
fit_lm <- lm(loss ~ exercise + food,
             data = df)

summary(fit_lm)
```

$$
\hat{Y} = 6 + 2X_{exercise} -0.5X_{food}
$$




Figure 3.5, page 54

```{r}
effects::Effect(focal.predictors = c("exercise", "food"),
                mod = fit_lm,
                xlevels = list(exercise = c(0, 2, 4))) %>% 
  data.frame %>% 
  dplyr::mutate(exercise = factor(exercise)) %>%
  ggplot(aes(x = food,
             y = fit,
             fill = exercise)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se),
              alpha = .2) +
  geom_line(aes(color = exercise)) +
  theme_bw() +
  labs(x = "Food Intake, 100s of calories above recommended",
       y = "Weight Loss, 100s of grams",
       fill  = "Exercise Frequency, average weekly hours: ",
       color = "Exercise Frequency, average weekly hours: ") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) 
```

Figrue 3.6, page 54

```{r}
effects::Effect(focal.predictors = c("exercise", "food"),
                mod = fit_lm,
                xlevels = list(food = c(0, 4,  8))) %>% 
  data.frame %>% 
  dplyr::mutate(food = factor(food)) %>%
  ggplot(aes(x = exercise,
             y = fit,
             fill = food)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se),
              alpha = .2) +
  geom_line(aes(color = food)) +
  theme_bw() +
  labs(x = "Food Intake, 100s of calories above recommended",
       y = "Exercise Frequency, average weekly hours",
       fill  = "Weight Loss, 100s of grams: ",
       color = "Weight Loss, 100s of grams: ") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) 
```



```{r}
anova(fit_lm)
```


```{r}
jtools::summ(fit_lm, 
             conf = TRUE,
             part.corr = TRUE)
```

```{r}
fit_lm_Z <- lm(scale(loss) ~ scale(exercise) + scale(food),
             data = df)
```

```{r}
summary(fit_lm_Z)
```


```{r, results='asis'}
texreg::knitreg(list(fit_lm, fit_lm_Z),
                single.row = TRUE)
```
