# Logistic Regression - Ex: Maternal Risk Factor for Low Birth Weight Delivery

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

```{r, message=FALSE, error=FALSE}
library(tidyverse)
library(haven)        # read in SPSS dataset
library(furniture)    # nice table1() descriptives
library(stargazer)    # display nice tables: summary & regression
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail
library(car)          # Companion to Applied Regression
library(sjPlot)       # Quick plots and tables for models
library(pscl)         # psudo R-squared function
library(glue)         # Interpreted String Literals 
```

## Background

More complex example demonstrating modeling decisions

Another set of data from a study investigating predictors of low birth weight

* `id` infant's unique identification number

Dependent variable (DV) or outcome    

* `low` Low birth weight (outcome) (0 = birth weight >2500 g (normal), 1 = birth weight < 2500 g (low))
* `bwt` actual infant birth weight in grams


Indepdentend variables (IV) or predictors

* `age` Age of mother, in years
* `lwt` Mother's weight at last menstrual period, in pounds
* `race` Race: 1 = White, 2 = Black, 3 = Other
* `smoke` Smoking status during pregnancy:1 = Yes, 0 = No
* `ptl` History of premature labor: 0 = None, 1 = One, 2 = two, 3 = three
* `ht` History of hypertension: 1 = Yes, 0 = No
* `ui` Uterine irritability: 1 = Yes, 0 = No
* `ftv` Number of physician visits in 1st trimester: 0 = None, 1 = One, ... 6 = six


### Raw Dataset

The data is saved in a text file (.txt) without any labels.

```{r lowbwtdata}
lowbwt_raw <- read.table("https://raw.githubusercontent.com/CEHS-research/data/master/Regression/lowbwt.txt", 
                         header = TRUE, 
                         sep = "", 
                         na.strings = "NA", 
                         dec = ".", 
                         strip.white = TRUE)

tibble::glimpse(lowbwt_raw)
```



### Declare Factors


```{r}
lowbwt_clean <- lowbwt_raw %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(low = factor(low,
                             levels = c(0, 1),
                             labels = c("birth weight >2500 g (normal)",
                                        "birth weight < 2500 g (low)"))) %>% 
  dplyr::mutate(race = factor(race,
                              levels = 1:3,
                              labels = c("White", 
                                         "Black", 
                                         "Other"))) %>% 
  dplyr::mutate(ptl_any = as.numeric(ptl > 0)) %>%         # colapse into 0 = none vs. 1 = at least one
  dplyr::mutate(ptl = factor(ptl)) %>%                     # declare the number of pre-term labors to be a factor: 0, 1, 2, 3
  dplyr::mutate_at(vars(smoke, ht, ui, ptl_any),           # declare all there variables to be factors with the same two levels
                   factor,
                   levels = 0:1,
                   labels = c("No", "Yes")) 
```


Display the structure of the 'clean' version of the dataset

```{r}
str(lowbwt_clean)
```





## Exploratory Data Analysis

```{r}
# Knit to Website: output = "html" 
# Knit to PDF:     output = "latex2"
# View on Screen:  output = ""text", or "markdown", "html"

lowbwt_clean %>% 
  furniture::table1("Age, years" = age, 
                    "Weight, pounds" = lwt, 
                    "Race" = race, 
                    "Smoking During pregnancy" = smoke, 
                    "History of Premature Labor, any" = ptl_any, 
                    "History of Premature Labor, number" = ptl, 
                    "History of Hypertension" = ht, 
                    "Uterince Irritability" = ui, 
                    "1st Tri Dr Visits" = ftv, 
                    splitby = ~ low,
                    test = TRUE,
                    output = "html")
```






## Logistic Regression - Simple, un-adjusted

```{r}
low1.age   <- glm(low ~ age,     family=binomial(logit), data=lowbwt_clean)
low1.lwt   <- glm(low ~ lwt,     family=binomial(logit), data=lowbwt_clean)
low1.race  <- glm(low ~ race,    family=binomial(logit), data=lowbwt_clean)
low1.smoke <- glm(low ~ smoke,   family=binomial(logit), data=lowbwt_clean)
low1.ptl   <- glm(low ~ ptl_any, family=binomial(logit), data=lowbwt_clean)
low1.ht    <- glm(low ~ ht,      family=binomial(logit), data=lowbwt_clean)
low1.ui    <- glm(low ~ ui,      family=binomial(logit), data=lowbwt_clean)
low1.ftv   <- glm(low ~ ftv,     family=binomial(logit), data=lowbwt_clean)
```

Note: the parameter estimates here are for the LOGIT scale, not the odds ration (OR) or even the probability.

```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(low1.age, low1.lwt, low1.race, low1.smoke),
                custom.model.names = c("Age", "Weight", "Race", "Smoker"),
                caption = "Simple, Unadjusted Logistic Regression: Models 1-4",
                caption.above = TRUE)
```


```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(low1.ptl, low1.ht, low1.ui, low1.ftv),
                custom.model.names = c("Pre-Labor", "Hypertension", "Uterine", "Visits"),
                caption = "Simple, Unadjusted Logistic Regression: Models 5-8",
                caption.above = TRUE)
```

## Logistic Regression - Multivariate, with Main Effects Only

Main-effects multiple logistic regression model

```{r}
low1_1 <- glm(low ~ age + lwt + race + smoke + ptl_any + ht + ui,
              family = binomial(logit), 
              data = lowbwt_clean)

summary(low1_1)
```


 



## Logistic Regression - Multivariate, with Interactions

Before removing non-significant main effects, test plausible interactions

Try interactions between age and lwt, age and smoke, lwt and smoke,  1 at a time

### Age and Weight

```{r}
low1_2 <- glm(low ~ age + lwt + race + smoke + ptl_any + ht + ui + age:lwt,
                 family = binomial(logit), 
                 data = lowbwt_clean)

summary(low1_2)
```

#### Compare Model Fits vs. Likelihood Ratio Test

```{r}
anova(low1_1, low1_2, test = 'Chi')
```

#### Type II Analysis of Deviance Table

```{r}
Anova(low1_2, test = 'LR') 
```

#### Type III Analysis of Deviance Table

```{r}
Anova(low1_2, test = 'LR', type = 'III') 
```


### Age and Smoking


```{r}
low1_3 <- glm(low ~ age + lwt + race + smoke + ptl_any + ht + ui + age:smoke,
                 family = binomial(logit), 
                 data = lowbwt_clean)

summary(low1_3)
```

#### Compare Model Fits vs. Likelihood Ratio Test

```{r}
anova(low1_1, low1_3, test = 'Chi')
```



### Weight and Smoking


```{r}
low1_4 <- glm(low ~ age + lwt + race + smoke + ptl_any + ht + ui + lwt:smoke,
                 family = binomial(logit), 
                 data = lowbwt_clean)

summary(low1_4)
```

#### Compare Model Fits vs. Likelihood Ratio Test

```{r}
anova(low1_1, low1_4, test = 'Chi')
```

## Logistic Regression - Multivariate, Simplify

No interactions are  significant
Remove non-significant main effects


### Remove the least significant perdictor: `ui`

```{r}
low1_5 <- glm(low ~ age + lwt + race + smoke + ptl_any + ht,
                 family = binomial(logit), 
                 data = lowbwt_clean)

summary(low1_5)
```


## Logistic Regression - Multivariate, Final Model

Since the mother's age is theoretically a meaningful variable, it should probably be retained.

Revise so that age is interpreted in 5-year and lwt in 20 lb increments and the intercept has meaning.



```{r}
low1_6 <- glm(low ~ I((age - 20)/5) + I((lwt - 125)/20) + race + smoke + ptl_any + ht,
              family = binomial(logit), 
              data = lowbwt_clean)

summary(low1_6)
```


### Several $R^2$ measures with the `pscl::pR2()` function 

```{r}
pscl::pR2(low1_6)
```


### Parameter Estiamtes Table

#### Using `texreg::screenreg()`

Default: parameters are in terms of the 'logit' or log odds ratio

```{r, include=FALSE}
texreg::screenreg(low1_6)
```

```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(low1_6)
```



The `texreg` package uses an intermediate function called `extract()` to extract information for the model and then put it in the right places in the table.  We can invervene by writing our own `extract_exp()` function to use instead.

```{r}
extract_exp <- function(fit_glm){
  beta   = coef(fit_glm)    
  betaci = confint(fit_glm) 
  fit_glm_exp        = texreg::extract(fit_glm)
  fit_glm_exp@coef   = exp(beta)
  fit_glm_exp@ci.low = exp(betaci[, 1])
  fit_glm_exp@ci.up  = exp(betaci[, 2])
  return(fit_glm_exp)
}
```

```{r, include=FALSE}
texreg::screenreg(extract_exp(low1_6),
                  custom.coef.names = c("BL: 125 lb, 20 yr old White Mother",
                                        "Additional 5 years older",
                                        "Additional 20 lbs pre-pregnancy",
                                        "Race: Black vs. White",
                                        "Race: Other vs. White",
                                        "Smoking During pregnancy",
                                        "History of Any Premature Labor",
                                        "History of Hypertension"),
                  custom.model.names = "OR, Low Birth Weight",
                  single.row = TRUE,
                  custom.note = "* The value of '1' is outside the confidence interval for the OR")
```

```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(extract_exp(low1_6),
                custom.coef.names = c("BL: 125 lb, 20 yr old White Mother",
                                      "Additional 5 years older",
                                      "Additional 20 lbs pre-pregnancy",
                                      "Race: Black vs. White",
                                      "Race: Other vs. White",
                                      "Smoking During pregnancy",
                                      "History of Any Premature Labor",
                                      "History of Hypertension"),
                custom.model.names = "OR, Low Birth Weight",
                single.row = TRUE,
                custom.note = "* The value of '1' is outside the confidence interval for the OR")
```




#### Using `sjPlot::tab_model()`

Parameters Exponentiated:

```{r, results = "asis"}
sjPlot::tab_model(low1_6,
                  emph.p = TRUE,
                  pred.labels = c("BL: 125 lb, 20 yr old White Mother",
                                  "Additional 5 years old",
                                  "Additional 20 lbs pre-pregnancy",
                                  "Race: Black vs. White",
                                  "Race: Other vs. White",
                                  "Smoking During pregnancy",
                                  "History of Any Premature Labor",
                                  "History of Hypertension")) 
```


### Marginal Model Plot

#### Focus on: Mother's Age, weight, and race

```{r, fig.width=8, fig.height=6}
effects::Effect(focal.predictors = c("age", "lwt", "race"),
                mod = low1_6,
                xlevels = list(age = c(20, 30, 40),
                               lwt = seq(from = 80, to = 250, by = 5))) %>% 
  data.frame() %>% 
  dplyr::mutate(age_labels = glue("Mother Age: {age}")) %>% 
  ggplot(aes(x = lwt,
             y = fit)) +
  geom_line(aes(color = race,
                linetype = race),
            size = 1) +
  theme_bw() +
  facet_grid(.~ age_labels) +
  labs(title = "Risk of Low Birth Weight",
       subtitle = "Illustates risk given mother is a non-smoker, without a history of pre-term labor or hypertension",
       x = "Mother's Weight Pre-Pregnancy, pounds",
       y = "Predicted Probability\nBaby has Low Birth Weight (< 2500 grams)",
       color    = "Mother's Race",
       linetype = "Mother's Race") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  scale_linetype_manual(values = c("longdash", "dotted", "solid")) +
  scale_color_manual(values = c( "coral2", "dodger blue", "gray50"))
```

#### Focus on: Mother's weight and smothing status during pregnancy, as well as history of any per-term labor and hypertension

```{r, fig.width=8, fig.height=6}
effects::Effect(focal.predictors = c("lwt", "smoke", "ptl_any", "ht"),
                fixed.predictors = list(age = 20),
                mod = low1_6,
                xlevels = list(lwt = seq(from = 80, to = 250, by = 5))) %>% 
  data.frame() %>% 
  dplyr::mutate(smoke = forcats::fct_rev(smoke)) %>% 
  dplyr::mutate(ptl_any_labels = glue("History of Preterm Labor: {ptl_any}")) %>% 
  dplyr::mutate(ht_labels = glue("History of Hypertension: {ht}") %>% forcats::fct_rev()) %>% 
  ggplot(aes(x = lwt,
             y = fit)) +
  geom_line(aes(color = smoke,
                linetype = smoke),
            size = 1) +
  theme_bw() +
  facet_grid(ht_labels ~ ptl_any_labels) +
  labs(title = "Risk of Low Birth Weight",
       subtitle = "Illustates risk given the mother is 20 years old and white",
       x = "Mother's Weight Pre-Pregnancy, pounds",
       y = "Predicted Probability\nBaby has Low Birth Weight (< 2500 grams)",
       color    = "Mother Smoked",
       linetype = "Mother Smoked") +
  theme(legend.position = c(1, .5),
        legend.justification = c(1.1, 1.15),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm")) +
  scale_linetype_manual(values = c("longdash", "solid")) +
  scale_color_manual(values = c( "coral2", "dodger blue"))
```

