# Extensions of Logistic Regression - Ex: Canadian Women's Labour-Force Participation

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

```{r, message=FALSE, error=FALSE}
library(tidyverse)
library(haven)        # read in SPSS dataset
library(furniture)    # nice table1() descriptives
library(stargazer)    # display nice tables: summary & regression
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail
library(car)          # Companion to Applied Regression
library(sjPlot)       # Quick plots and tables for models
library(car)          # Companion to Applied Regression (a text book - includes datasets)
library(MASS)         # Support Functions and Datasets
library(nnet)         #  Multinomial Log-Linear Models
```

## Background

The `Womenlf` data frame has 263 rows and 4 columns. The data are from a 1977 survey of the Canadian population.

Dependent variable (DV) or outcome    

* `partic` Labour-Force Participation, a factor with levels: 
    + `fulltime` Working full-time    
    + `not.work` Not working outside the home    
    + `parttime` Working part-time    


Indepdentend variables (IV) or predictors

* `hincome` Husband's income, in $1000's    

* `children` Presence of children in the household, a factor with levels:     
    + `absent` no children in the home   
    + `present` at least one child at home

* `region` A factor with levels:     
    + `Atlantic` Atlantic Canada
    + `BC` British Columbia
    + `Ontario` 
    + `Prairie` Prairie provinces
    + `Quebec`


### Raw Dataset

The data is included in the `carData` package which installs and loads with the `car` package.  

```{r}
data(Womenlf, package = "carData")  # load the internal data

tibble::glimpse(Womenlf)            # glimpse a bit of the data
```

```{r}
Womenlf %>% 
  dplyr:: filter(row_number() %in% sample(1:nrow(.), size = 10))  # select a random sample of 10 rows
```


Notice the order of the factor levels, especially for the `partic` factor

```{r}
str(Womenlf)  # view the structure of the data
```


We can view the order of the factors levels 

```{r}
Womenlf$partic %>% levels()  # view the levels (in order) of the variable
```


### Declare Factors

```{r}
Womenlf_clean <- Womenlf %>% 
  dplyr::mutate(working_ord = partic %>% 
                  forcats::fct_recode("Full Time"  = "fulltime",
                                      "Not at All" = "not.work",
                                      "Part Time"  = "parttime") %>% 
                  factor(levels = c("Not at All", 
                                    "Part Time", 
                                    "Full Time"))) %>% 
  dplyr::mutate(working_any  = dplyr::case_when(partic %in% c("fulltime", "parttime") ~ "At Least Part Time",
                                                partic == "not.work" ~ "Not at All") %>% 
                  factor(levels = c("Not at All",                                
                                    "At Least Part Time"))) %>%     
  dplyr::mutate(working_full = dplyr::case_when(partic == "fulltime" ~ "Full Time",    
                                                partic %in% c("not.work", "parttime") ~ "Less Than Full Time")%>% 
                  factor(levels = c("Less Than Full Time",
                                    "Full Time"))) %>%         
  dplyr::mutate(working_type = dplyr::case_when(partic == "fulltime" ~ "Full Time",        
                                                partic == "parttime" ~ "Part Time")%>%      
                  factor(levels = c("Part Time",                                
                                    "Full Time")))                            
```


Display the structure of the 'clean' version of the dataset

```{r}
str(Womenlf_clean)  # view the structure of the data 
```

## Exploratory Data Analysis

Three versions of the outcome

```{r}
Womenlf_clean %>%
  furniture::table1(working_ord, working_any, 
                    working_full, working_type,
                    na.rm = FALSE,              # do NOT restrict to complete cases!!!
                    output = "markdown")
```


Other Predisctors, univariate


```{r}
Womenlf_clean %>%
  furniture::table1("Husband's Income, $1000's" = hincome, 
                    "Children In the Home" = children, 
                    "Region of Canada" = region,
                    output = "markdown")
```

```{r}
Womenlf_clean %>%
  furniture::table1("Husband's Income, $1000's" = hincome, 
                    "Children In the Home" = children, 
                    "Region of Canada" = region,
                    splitby = ~ working_ord, 
                    row_wise = TRUE,                        # show row %s rather than default column %s
                    test = TRUE,
                    total = TRUE,
                    output = "markdown")
```

### Husband's Income

```{r}
Womenlf_clean %>% 
  ggplot(aes(hincome,
             fill = working_ord)) +
  geom_density(alpha = .3)
```

```{r}
Womenlf_clean %>% 
  ggplot(aes(x = working_ord,
             y = hincome)) +
  geom_jitter(position=position_jitter(0.2))  +
    stat_summary(fun.y = mean, 
                 geom = "errorbar", 
                 aes(ymax = ..y.., 
                     ymin = ..y..),
                 width = .75, 
                 color = "red",
                 size = 1)
```

```{r}
Womenlf_clean %>% 
  ggplot(aes(hincome,
             x = working_ord,
             fill = working_ord)) +
  geom_boxplot(alpha = .3)
```

```{r}
Womenlf_clean %>% 
  ggplot(aes(hincome,
             x = working_ord,
             fill = working_ord)) +
  geom_violin(alpha = .3) + 
  stat_summary(fun = mean, 
               geom = "errorbar", 
               aes(ymax = ..y.., 
                   ymin = ..y..),
               width = .75, 
               color = "red",
               size = 1)
```

### Children in Home


```{r}
Womenlf_clean %>% 
  ggplot(aes(x = children,
             fill = working_ord)) +
  geom_bar()
```


```{r}
Womenlf_clean %>% 
  ggplot(aes(x = children,
             fill = working_ord %>% fct_rev)) +
  geom_bar(position="fill") +
  labs(x = "Children in the Home",
       y = "Proportion of Women",
       fill = "Working") +
  theme_bw() +
  scale_fill_manual(values = c("gray25", "gray50", "gray75"))
```

### Region of Canada

```{r}
Womenlf_clean %>% 
  ggplot(aes(x = region,
             fill = working_ord)) +
  geom_bar()
```

```{r}
Womenlf_clean %>% 
  ggplot(aes(x = region,
             fill = working_ord %>% fct_rev)) +
  geom_bar(position="fill") +
  labs(x = "Region of Canada",
       y = "Proportion of Women",
       fill = "Working") +
  theme_bw() +
  scale_fill_manual(values = c("gray25", "gray50", "gray75"))
```

## Hierarchical (nested) Logistic Regression

For an $m-$category polytomy dependent variable is respecified as a series of $m – 1$ nested dichotomies.  A single or combined levels of outcome compared to another single or combination of levels.  Then they are analyzed using a series of binary logistic regressions, such that:

* Dichotomies selected based on theory
* Avoid redundancy
* Similar to contrast coding, but for outcome

For this dataset example, the outcome (`partic`) has $3-$categories, so we will investigate TWO nested dichotomies

1. outcome = `working_any`
2. outcome = `working_type`

### Role of Predictors on ANY working

Fit a regular logistic model with all three predictors regressed on the binary indicator for any working.  Use the `glm()` function in the base $R$ `stats` package.

```{r}
fit_glm_1 <- glm(working_any ~ hincome + children + region,
                 data = Womenlf_clean,
                 family = binomial(link = "logit"))

summary(fit_glm_1)
```

Check if `region` is statistically significant with the `drop1()` function from the base $R$ `stats` package.  This may be done with a Likelihood Ratio Test (`test = "LRT"`, which is the same as `test = "Chisq"` for `glm` models).

```{r}
drop1(fit_glm_1, test = "LRT")
```

Since the region doesn't have exhibit any effect on odds a women is in the labor force, remove that predictor in the model to simplify to a 'best' final model.  Also, center husband's income at a value near the mean so the intercept has meaning.

```{r}
fit_glm_2 <- glm(working_any ~ I(hincome - 14) + children,
                 data = Womenlf_clean,
                 family = binomial(link = "logit"))

summary(fit_glm_2)
```

The `texreg` package uses an intermediate function called `extract()` to extract information for the model and then put it in the right places in the table.  We can invervene by writing our own `extract_exp()` function to use instead.

```{r}
extract_exp <- function(fit_glm){
  beta   = coef(fit_glm)    
  betaci = confint(fit_glm) 
  fit_glm_exp        = texreg::extract(fit_glm)
  fit_glm_exp@coef   = exp(beta)
  fit_glm_exp@ci.low = exp(betaci[, 1])
  fit_glm_exp@ci.up  = exp(betaci[, 2])
  return(fit_glm_exp)
}
```



```{r, results='asis'}
texreg::knitreg(extract_glm_exp(fit_glm_2),
                custom.coef.names = c("BL: No children, Husband Earns $14,000/yr",
                                      "Husband's Income, $1000's",
                                      "Children in the Home"),
                custom.model.names = "OR, Women is in the Workforce at All",
                single.row = TRUE,
                custom.note = "* The value of '1' is outside the confidence interval for the OR")
```


**Interpretation: **

* Among women without children in the home and a husband making $14,000 annually, there is about a 2:1 odds she is in the workforce.

* For each additional thousand dollars the husband makes, the odds ratio decreases by about 4 percent.

* If there are children in the home, the odds of being in the workforce is nearly a fifth as large.

```{r}
interactions::interact_plot(model = fit_glm_2,
                            pred = hincome,
                            modx = children,
                            interval = TRUE)
```



```{r}
interactions::interact_plot(model = fit_glm_2,
                            pred = hincome,
                            modx = children,
                            interval = TRUE,
                            x.label = "Husband's Income, in $1,000's",
                            y.label = "Predicted Probability of\nWomen Being in the Workforce",
                            legend.main = "Children in the Home:",
                            modx.labels = c("Absent", "Present"),
                            colors = rep("black", 2)) +
  geom_vline(xintercept = 14, color = "gray25") +   # reference line for intercept
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"))
```



```{r}
effects::allEffects(fit_glm_2)
```

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_glm_2) %>% 
  data.frame() %>% 
  ggplot(aes(x = hincome,
             y = fit,
             color = children,
             linetype = children)) +
  geom_vline(xintercept = 14, color = "gray25") +   # reference line for intercept
  geom_line(size = 1) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability of\nWomen Being in the Workforce",
       color    = "Children in\nthe Home:",
       linetype = "Children in\nthe Home:") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black")) +
  coord_cartesian(ylim = c(0, 1))
```




### Role of Predictors on TYPE of work


Fit a regular logistic model with all three predictors regressed on the binary indicator for level/type of working.

```{r}
fit_glm_3 <- glm(working_type ~ hincome + children + region,
                 data = Womenlf_clean,
                 family = binomial(link = "logit"))

summary(fit_glm_3)
```


Check if `region` is statistically significant with the `drop1()` function from the base $R$ `stats` package.  This may be done with a Likelihood Ratio Test (`test = "LRT"`, which is the same as `test = "Chisq"` for `glm` models).

```{r}
drop1(fit_glm_3, test = "LRT")
```

Since the region doesn't have exhibit any effect on odds a working women is in the labor force full time, remove that predictor in the model to simplify to a 'best' final model.  Also, center husband's income at a value near the mean so the intercept has meaning.

```{r}
fit_glm_4 <- glm(working_type ~ I(hincome - 14) + children,
                 data = Womenlf_clean,
                 family = binomial(link = "logit"))

summary(fit_glm_4)
```

The `texreg` package uses an intermediate function called `extract()` to extract information for the model and then put it in the right places in the table.  We can invervene by writing our own `extract_exp()` function to use instead.




```{r, results='asis'}
texreg::knitreg(list(extract_glm_exp(fit_glm_2), 
                     extract_glm_exp(fit_glm_4)),
                custom.coef.names = c("BL: No children, Husband Earns $14,000/yr",
                                      "Husband's Income, $1000's",
                                      "Children in the Home"),
                custom.model.names = c("Working at All",
                                       "Full vs. Part-Time"),
                single.row = TRUE,
                custom.note = "* The value of '1' is outside the confidence interval for the OR")
```


**Interpretation: **

* Among working women without children in the home and a husband making $14,000 annually, there is more than 7:1 odds she is working full time verses part time.

* For each additional thousand dollars the husband makes, the odds ratio decreases by about 10 percent.

* If there are children in the home, the odds of being in the workforce is drastically reduced.



```{r}
interactions::interact_plot(model = fit_glm_4,
                            pred = hincome,
                            modx = children,
                            interval = TRUE)
```



```{r}
interactions::interact_plot(model = fit_glm_4,
                            pred = hincome,
                            modx = children,
                            interval = TRUE,
                            x.label = "Husband's Income, in $1,000's",
                            y.label = "Predicted Probability of\nWomen Being in the Workforce Full Time\nif they are working",
                            legend.main = "Children in the Home:",
                            modx.labels = c("Absent", "Present"),
                            colors = rep("black", 2)) +
  geom_vline(xintercept = 14, color = "gray25") +   # reference line for intercept
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"))
```


```{r}
effects::allEffects(fit_glm_4)
```

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_glm_4) %>% 
  data.frame() %>% 
  ggplot(aes(x = hincome,
             y = fit,
             color = children,
             linetype = children)) +
  geom_vline(xintercept = 14, color = "gray25") +   # reference line for intercept
  geom_line(size = 1) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability of\nWomen Being in the Workforce Full Time\nif they are working",
       color    = "Children in\nthe Home:",
       linetype = "Children in\nthe Home:") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black")) +
  coord_cartesian(ylim = c(0, 1))
```

## Multinomial (nominal) Logistic Regression

Multinomial Logistic Regression fits a single model by specifing a reference level of the outcome and comparing each additional level to it.  In our case we will choose `not working` as the reference category adn get a set of parameter estimates (betas) for each of the two options `part time` and `full time`.

### Fit Model 1: main effects only

Use `multinom()` function in the base $R$ `nnet` package.  You will also need the `MASS` and $R$ package (only to compute MLEs).  Make sure to remove cases with missing data on predictors before modeling or use the `na.action = na.omit` optin in the `multinom()` model command.

```{r}
fit_multnom_1 <- nnet::multinom(working_ord ~ I(hincome - 14) + children + region,
                                data = Womenlf_clean)

summary(fit_multnom_1,
        corr = FALSE,
        wald = TRUE)
```


### Fit Model 2: only significant predictors

Reduce the model by removing the `region` variable.

```{r}
fit_multnom_2 <- nnet::multinom(working_ord ~ I(hincome - 14) + children,
                                data = Womenlf_clean)

summary(fit_multnom_2,
        corr = FALSE,
        wald = TRUE)
```


### Compre model fit

Check if we need to keep the `region` variable in our model.

```{r}
anova(fit_multnom_1, fit_multnom_2)
```

```{r}
performance::compare_performance(fit_multnom_1, fit_multnom_2, rank = TRUE)
```


### Extract parameters


#### Logit Scale

Here is one way to extract the parameter estimates, but recall they are in terms of the **logit** or log-odds, not probability.

```{r}
broom::tidy(fit_multnom_2) %>% 
  dplyr::mutate(p.value = round(p.value, 4))
```


#### Odds-Ratio Scale

The `effects::allEffects()` function provides probability estimates for each outcome level for different levels of the predictors.

```{r}
effects::allEffects(fit_multnom_2)
```

### Tabulate parameters

The `texreg` package know how to handle this type of model and displays the parameters estimates in two separate columns. 

```{r}
texreg::extract(fit_multnom_2)
```


```{r, results='asis'}
texreg::knitreg(fit_multnom_2,
                custom.model.name = c("b (SE)"),
                custom.coef.map = list("Part Time: (Intercept)"     = "PT-BL: No children, Husband Earns $14,000/yr",
                                       "Part Time: I(hincome - 14)" = "PT-Husband's Income, in $1,000's",
                                       "Part Time: childrenpresent" = "PT-Children in the Home",
                                       "Full Time: (Intercept)"     = "FT-BL: No children, Husband Earns $14,000/yr",
                                       "Full Time: I(hincome - 14)" = "FT-Husband's Income, in $1,000's",
                                       "Full Time: childrenpresent" = "FT-Children in the Home"),
                groups = list("Part Time" = 1:3,
                              "Full Time" = 4:6),
                single.row = TRUE)
```


```{r}
fit_multnom_2 %>% coef() 
```


```{r}
fit_multnom_2 %>% coef() %>% exp()
```

```{r}
fit_multnom_2 %>% confint() %>% exp()
```




### Plot Predicted Probabilities

> NOTE:  I'm not sure how to use the `interactions::interact_plot()` function with multinomial models.

#### Manually Compute

* estimates for probabilities and the associated 95% confidence intervals


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_multnom_2)
```

#### Wrange 

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_multnom_2) %>% 
  data.frame() 
```


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_multnom_2) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time)
```


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_multnom_2) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit")
                      
```

#### Plot, version 1


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_multnom_2) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(work_level = work_level %>% 
                  factor() %>% 
                  forcats::fct_recode("Not at All" = "none",
                                      "Part Time"  = "part",
                                      "Full Time"  = "full") %>% 
                  forcats::fct_rev()) %>%
  ggplot(aes(x = hincome,
             y = fit)) +
  geom_vline(xintercept = 14,  color = "gray50") +  # reference line for intercept
  geom_hline(yintercept = 0.5, color = "gray50") + # 50% chance line for reference
  geom_ribbon(aes(ymin = lower,
                  ymax = upper,
                  fill = work_level),
              alpha = .3) + 
  geom_line(aes(color = work_level,
                linetype = work_level),
            size = 1) +
  facet_grid(. ~ children, labeller = label_both) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability",
       color    = "Woman Works:",
       fill     = "Woman Works:",
       linetype = "Woman Works:") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_linetype_manual(values = c("solid", "longdash", "dotted"))
```

#### Plot, version 2

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_multnom_2) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(work_level = factor(work_level,
                                    levels = c("none", "part", "full"),
                                    labels = c("Not at All",
                                               "Part Time",
                                               "Full Time"))) %>% 
  ggplot(aes(x = hincome,
             y = fit)) +
  geom_vline(xintercept = 14,  color = "gray50") +  # reference line for intercept
  geom_hline(yintercept = 0.5, color = "gray50") + # 50% chance line for reference
  geom_ribbon(aes(ymin = lower,
                  ymax = upper,
                  fill = children),
              alpha = .3) + 
  geom_line(aes(color    = children,
                linetype = children),
            size = 1) +
  facet_grid(. ~ work_level) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability",
       color    = "Children in the Home:",
       fill     = "Children in the Home:",
       linetype = "Children in the Home:") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_fill_manual(values = c("dodgerblue", "coral3"))+
  scale_color_manual(values = c("dodgerblue", "coral3"))
```


### Interpretation

* Among women without children in the home and a husband making $14,000 annually, there is about 1:4 odds she is working part time verses not at all. and a 1.8:1 odds she is working full time.

* For each additional thousand dollars the husband makes, the odds ratio decreases by about 10 percent that she is working full time, yet stay the same that she works part time.

* If there are children in the home, the odds of working part time increase by 2 percent and there is a very unlikely change she works full time.




## Proportional-odds (ordinal) Logistic Regression

This type of logisit regression model forces the predictors to have similar relationship with the outcome (slopes), but different means (intercepts).  This is called the proportional odds assumption.

### Fit the Model

Use `polr()` function in the base $R$ `MASS` package.  While outcome variable (dependent variable, "Y") may be a regular `factor`, it is preferable to specify it as an `ordered` factor.

```{r}
fit_polr_1 <- MASS::polr(working_ord ~ hincome + children,
                         data = Womenlf_clean)

summary(fit_polr_1)
```

### Extract Parameters

#### Logit Scale


```{r}
fit_polr_1$zeta
```



```{r}
fit_polr_1 %>% confint()
```

#### Odds-Ratio Scale

```{r}
fit_polr_1 %>% coef() %>% exp()
```


```{r}
fit_polr_1 %>% confint() %>% exp()
```

#### Predicted Probabilities

```{r}
effects::allEffects(fit_polr_1)
```

### Tabulate parameters


```{r, results='asis'}
texreg::knitreg(fit_polr_1,
                custom.model.name = "Proportional Odds",
                custom.coef.map = list("hincome" = "Husband's Income, $1000's",
                                       "childrenpresent" = "Children in the Home",
                                       "Not at All|Part Time" = "Not at All vs.Part Time",
                                       "Part Time|Full Time"  = "Part Time vs.Full Time"),
                groups = list("Coefficents" = 1:2,
                              "Intercepts" = 3:4),
                caption = "Proportional-odds (ordinal) Logistic Regression",
                caption.above = TRUE,
                single.row = TRUE)
```


### Plot Predicted Probabilities

#### Manually Compute

* estimates for probabilities and the associated 95% confidence intervals


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_polr_1)
```

#### Wrangle

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_polr_1) %>% 
  data.frame()
```


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_polr_1) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time)
```
```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = c(20, 30, 40)),
                mod = fit_polr_1) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit")
                      
```

#### Plot, version 1


```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_polr_1) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(work_level = work_level %>% 
                  factor() %>% 
                  forcats::fct_recode("Not at All" = "none",
                                      "Part Time"  = "part",
                                      "Full Time"  = "full") %>% 
                  forcats::fct_rev()) %>%
  ggplot(aes(x = hincome,
             y = fit)) +
  geom_vline(xintercept = 14,  color = "gray50") +  # reference line for intercept
  geom_hline(yintercept = 0.5, color = "gray50") + # 50% chance line for reference
  geom_ribbon(aes(ymin = lower,
                  ymax = upper,
                  fill = work_level),
              alpha = .3) + 
  geom_line(aes(color = work_level,
                linetype = work_level),
            size = 1) +
  facet_grid(. ~ children, labeller = label_both) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability",
       color    = "Woman Works:",
       fill     = "Woman Works:",
       linetype = "Woman Works:") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_linetype_manual(values = c("solid", "longdash", "dotted"))
```



#### Plot, version 2

```{r}
effects::Effect(focal.predictors = c("hincome", "children"),
                xlevels = list(hincome = seq(from = 1, to = 45, by = .1)),
                mod = fit_polr_1) %>% 
  data.frame() %>% 
  dplyr::select(hincome, children, 
                starts_with("prob"), 
                starts_with("L.prob"), 
                starts_with("U.prob")) %>% 
  dplyr::rename(est_fit_none = prob.Not.at.All,
                est_fit_part = prob.Part.Time,
                est_fit_full = prob.Full.Time,
                est_lower_none = L.prob.Not.at.All,
                est_lower_part = L.prob.Part.Time,
                est_lower_full = L.prob.Full.Time,
                est_upper_none = U.prob.Not.at.All,
                est_upper_part = U.prob.Part.Time,
                est_upper_full = U.prob.Full.Time) %>% 
  tidyr::pivot_longer(cols = starts_with("est"),
                      names_to = c(".value", "work_level"),
                      names_pattern = "est_(.*)_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(work_level = factor(work_level,
                                    levels = c("none", "part", "full"),
                                    labels = c("Not at All",
                                               "Part Time",
                                               "Full Time"))) %>% 
  ggplot(aes(x = hincome,
             y = fit)) +
  geom_vline(xintercept = 14,  color = "gray50") +  # reference line for intercept
  geom_hline(yintercept = 0.5, color = "gray50") + # 50% chance line for reference
  geom_ribbon(aes(ymin = lower,
                  ymax = upper,
                  fill = children),
              alpha = .3) + 
  geom_line(aes(color    = children,
                linetype = children),
            size = 1) +
  facet_grid(. ~ work_level) +
  theme_bw() +
  labs(x = "Husband's Income, in $1000's",
       y = "Predicted Probability",
       color    = "Children in the Home:",
       fill     = "Children in the Home:",
       linetype = "Children in the Home:") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_fill_manual(values = c("dodgerblue", "coral3"))+
  scale_color_manual(values = c("dodgerblue", "coral3"))
```


### Interpretation

* Among women without children in the home and a husband making $14,000 annually, there is a 26% chance she is not working, 21% change she is working part time and just over a 53% change she is working full time.

* For each additional thousand dollars the husband makes, the odds ratio decreases by about 5 percent that she is working part time vs not at all and 5% that she is working full time vs part time.

* If there are children in the home, the odds ratio of working part time vs not at all decreases by 86% and similartly the odds ratio fo working full time vs part time also decreases by 86%.





## Compare Model Fits: Multinomial vs. Ordinal

The multinomail and proportional-odds models aren't truely nested, so you can NOT conduct a Likelihood-Ratio Test (aka Deviance Difference Test) with the `anova()` command.  

You can use the `performance::compare_performance()` command to compare overal model performance via the Bayes factor (`BF`).


```{r}
performance::compare_performance(fit_multnom_2, fit_polr_1, rank = TRUE)
```

### Interpretation

The multinomial model looks to fit these data better than the proportional-odds (aka ordinal) logisic model.

