# Ordered Logistic Regression - Ex: Spaking


```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

```{r, message=FALSE, error=FALSE}
library(tidyverse)
library(haven)        # read in SPSS dataset
library(furniture)    # nice table1() descriptives
library(stargazer)    # display nice tables: summary & regression
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail
library(car)          # Companion to Applied Regression
library(sjPlot)       # Quick plots and tables for models
library(car)          # Companion to Applied Regression (a text book - includes datasets)
library(MASS)         # Support Functions and Datasets
library(nnet)         #  Multinomial Log-Linear Models
library(pscl)         # Political Science Computational Laboratory (ZIP)
```

## Background

This dataset comes from John Hoffman's textbook: Regression Models for Categorical, Count, and Related Variables: An Applied Approach (2004) [Amazon link, 2014 edition](https://www.amazon.com/Regression-Models-Categorical-Related-Variables/dp/0520289293/ref=sr_1_2?dchild=1&qid=1603172859&refinements=p_27%3ADr.+John+P.+Hoffmann&s=books&sr=1-2&text=Dr.+John+P.+Hoffmann)

> Chapter 4: Ordered Logistic and Probit Regression Models 

Dataset:  The following example uses the SPSS data set `gss.sav`. The dependent variable of interest is labeled `spanking`. 


> " The pertinent question (`spanking`) asks "Do you strongly agree, agree, disagree, or strongly disagree that it is sometimes necessary to discipline a child with a good, hard spanking?" The possible answers are coded as `1` = strongly agree, `2` = agree, `3` = disagree, and `4` = strongly disagree. A common hypothesis is that support for corporal punishment of children decreases at higher levels of education."



### Raw Dataset

```{r}
data_gss <- haven::read_spss("https://raw.githubusercontent.com/CEHS-research/data/master/Hoffmann_datasets/gss.sav") %>% 
  haven::as_factor()

data_gss %>% 
  dplyr::select(spanking, female, nonwhite, educate, income) %>% 
  dplyr::filter(!is.na(spanking)) %>%      # about 1/3 of participants are missing this
  head()
```

### Wrangle Data

```{r}
data_gss_model <- data_gss %>% 
  dplyr::mutate(spankingN = as.numeric(spanking)) %>%   # numeric version: 1, 2, 3, 4
  dplyr::mutate(polviewsN = as.numeric(polviews)) %>% 
  dplyr::filter(complete.cases(educate, spanking))      # only include complete cases
```



## Exploratory Data Analysis

### Entire Sample


```{r}
data_gss %>% 
  furniture::table1(spanking,
                    na.rm = FALSE,
                    output = "markdown",
                    caption = "Hoffmann's Example 4.1 Summary of the Spanking Variable")
```




```{r}
data_gss %>% 
  ggplot(aes(spanking)) +
  geom_bar() 
```



### By Education


```{r}
data_gss %>% 
  dplyr::group_by(forcats::fct_explicit_na(spanking)) %>% 
  furniture::table1("Educations, years" = educate,
                    "Education, factor" = factor(educate),
                    na.rm  = FALSE,
                    digits = 2,
                    output = "markdown")
```

### Spanking by Sex


```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>% 
  dplyr::select(female, spanking) %>% 
  table() %>% 
  addmargins()
```



```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>%  
  furniture::tableX(female, spanking,
                    type = "count")
```


```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>%  
  furniture::tableX(female, spanking,
                    type = "row_perc")
```

```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>%  
  furniture::tableX(female, spanking,
                    type = "col_perc")
```


```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>%  
  furniture::tableX(female, spanking,
                    type = "cell_perc")
```

```{r}
data_gss %>% 
  dplyr::filter(complete.cases(female, spanking)) %>%   
  dplyr::group_by(spanking) %>% 
  furniture::table1(female)
```




## Linear Regression

Linear regression is often ill-suited to fitting a likert rating, such as agreement.

### Visualization

```{r, fig.cap="Hoffmann's Figure 4.1 "}
data_gss_model %>% 
  ggplot(aes(x = educate,
             y = spankingN)) +                    
  geom_count() +                               # point size relative to over-plotting
  geom_smooth(method = "lm") +                 # add linear regression line (OLS)
  theme_bw() +
  labs(x = "Years of Formal Education",
       y = "Spanking")
```


### Fit the Model

```{r}
fit_lm <- lm(spankingN ~ educate,
             data = data_gss_model)

summary(fit_lm)
```


```{r}
anova(fit_lm)
```

### Tabulate Parameters

```{r, results='asis'}
texreg::knitreg(fit_lm,
                custom.model.name = "Linear Regression",
                caption = "Hoffmann's Example 4.2",
                caption.above = TRUE,
                single.row = TRUE,
                digits = 4)
```

### Model Fit and Variance Explained

```{r}
performance::performance(fit_lm)
```

```{r}
performance::r2(fit_lm)
```

### Residual Diagnostics

```{r, fig.cap="Hoffman's Figures 4.2 adn 4.3 Residual Diagnostics for a linear model on likery dependent variable - YUCK!"}
sjPlot::plot_model(fit_lm, type = "diag")
```


## Ordered Logistic Regression


```{r}
data_gss_model %>% 
  dplyr::group_by(forcats::fct_explicit_na(spanking)) %>% 
  furniture::table1("Sex" = female,
                    caption = "Hoffmann's Example 4.3 Crosstabulate DV with Sex",
                    na.rm  = FALSE,
                    digits = 2,
                    total = TRUE,
                    output = "markdown")
```

```{r}
data_gss_model %>% 
  furniture::tableX(female, spanking)
```




## Proportional-odds (ordinal) Logistic Regression

This type of logisit regression model forces the predictors to have similar relationship with the outcome (slopes), but different means (intercepts).  This is called the proportional odds assumption.

### Fit Model 1: Sex

Use `polr()` function in the base $R$ `MASS` package.  While outcome variable (dependent variable, "Y") may be a regular `factor`, it is preferable to specify it as an `ordered` factor.

```{r}
fit_polr_1 <- MASS::polr(spanking ~ female,
                         data = data_gss_model)

summary(fit_polr_1)
```





### Extract Parameters

#### Logit Scale


```{r}
fit_polr_1$zeta
```

```{r}
fit_polr_1 %>% coef()
```


```{r}
fit_polr_1 %>% confint()
```


#### Odds-Ratio Scale
```{r}
fit_polr_1$zeta %>% exp()
```




```{r}
fit_polr_1 %>% coef() %>% exp()
```


```{r}
fit_polr_1 %>% confint() %>% exp()
```


#### Predicted Probabilities

```{r}
effects::allEffects(fit_polr_1)
```

### Tabulate parameters


```{r, results='asis'}
texreg::knitreg(fit_polr_1,
                custom.model.name = c("b (SE)"),
                custom.coef.map = list("femalefemale"             = "Female vs. Male",
                                       "strongly agree|agree"     = "strongly agree|agree",
                                       "agree|disagree"           = "agree|disagree",
                                       "disagree|strongly disagree" = "disagree|strongly disagree"),
                groups = list("Predictors" = 1,
                              "Cut Values (i.e. threasholds)" = 2:4),
                caption = "Hoffmann's Example 4.4 Ordered Logistic Regression",
                caption.above = TRUE,
                single.row = TRUE,
                digits = 4)
```


### Predicted Probabilities

```{r}
ggeffects::ggeffect(model = fit_polr_1,
                    terms = c("female"))
```


```{r}
ggeffects::ggeffect(model = fit_polr_1,
                    terms = c("female")) %>% 
  dplyr::filter(x == "female")
```




### Plot Predicted Probabilities


```{r}
ggeffects::ggeffect(model = fit_polr_1,
                    terms = c("female")) %>%    # x-axis
  data.frame() %>% 
  ggplot(aes(x = x,
             y = predicted,
             group = response.level,
             color = response.level)) +
  geom_errorbar(aes(ymin = conf.low,
                    ymax = conf.high),
                width = .25) +
  geom_point(size = 4) +
  geom_line(aes(linetype = response.level)) 

```





```{r}
ggeffects::ggeffect(model = fit_polr_1,
                    terms = c("female")) %>%    # x-axis
  data.frame() %>% 
  dplyr::mutate(response.level = response.level %>% 
                  forcats::fct_reorder(predicted) %>% 
                  forcats::fct_rev()) %>% 
  ggplot(aes(x = x,
             y = predicted,
             group = response.level,
             color = response.level)) +
  geom_errorbar(aes(ymin = conf.low,
                    ymax = conf.high),
                width = .25) +
  geom_point(size = 4) +
  geom_line(aes(linetype = response.level)) +
  theme_bw() +
  labs(x = NULL,
       y = "Predicted Probability",
       color    = "Spanking:",
       shape    = "Spanking:",
       linetype = "Spanking:") +
  theme(legend.key.width = unit(2, "cm")) +
  scale_linetype_manual(values = c("solid", "longdash", "dotdash", "dotted")) +
  scale_shape_manual(values = c(0, 1, 2, 8))
```




### Model Fit and Variance Explained

```{r}
fit_polr_0 <- MASS::polr(spanking ~ 1,
                         data = data_gss_model)
```



```{r}
anova(fit_polr_1, fit_polr_0)
```


```{r}
performance::performance(fit_polr_1)
```

```{r}
performance::r2(fit_polr_1)
```

### Assumptions

#### Proportional Odds: Brant Test


The `poTest` function implements tests proposed by Brant (1990) for proportional odds for logistic models fit by the `polr()` function in the MASS package.

```{r}
# Hoffmann's Examle 4.5 (continued...)
car::poTest(fit_polr_1)
```

A significant test statistics provides evidence that the parallel regression assumption has been violated!



### Fit Model 2: Sex + Covars


```{r}
fit_polr_2 <- MASS::polr(spanking ~ female + educate + polviewsN,
                         data = data_gss_model)

summary(fit_polr_2)
```



### Extract Parameters

#### Logit Scale


```{r}
fit_polr_2$zeta
```

```{r}
fit_polr_2 %>% coef()
```


```{r}
fit_polr_2 %>% confint()
```


#### Odds-Ratio Scale
```{r}
fit_polr_2$zeta %>% exp()
```




```{r}
fit_polr_2 %>% coef() %>% exp()
```


```{r}
fit_polr_2 %>% confint() %>% exp()
```

### Tabulate parameters


```{r, results='asis'}
texreg::knitreg(fit_polr_2,
                custom.model.name = c("b (SE)"),
                custom.coef.map = list("femalefemale"             = "Female vs. Male",
                                       "educate"                  = "Years of Education",
                                       "polviewsN"                = "Level of Polytical Views",
                                       "strongly agree|agree"     = "strongly agree|agree",
                                       "agree|disagree"           = "agree|disagree",
                                       "disagree|strongly disagree" = "disagree|strongly disagree"),
                groups = list("Predictors" = 1:3,
                              "Cut Values" = 4:6),
                caption = "Hoffmann's Example 4.7 Ordered Logistic Regression",
                caption.above = TRUE,
                single.row = TRUE,
                digits = 4)
```

### Predicted Probabilities

> The `ggeffects` package computes estimated marginal means (predicted values) for the response, at the margin of specific values or levels from certain model terms, *i.e. it generates predictions by a model by holding the non-focal variables constant and varying the focal variable(s)*. 
 
 * `ggpredict()` uses `predict()` for generating predictions
     - factors: uses the reference level
     
 * `ggeffect()` computes marginal effects by internally calling `effects::Effect()`
     - factors: compute a kind of "average" value, which represents the proportions of each factor's category
     
 * `ggemmeans()` uses `emmeans::emmeans()`
     - factors: compute a kind of "average" value, which represents the proportions of each factor's category
     

Use `condition` to set a specific level for factors in `ggemmeans()`, so factors are not averaged over their categories, but held constant at a given level.


> `ggeffects::ggpredict()`
> Adjusted for:
> *   educate = 13.51  *The grand mean value*
> * polviewsN =  4.17  *The grand mean value*



```{r}
## Hoffmann's Example 4.8 (continues...approximated)
ggeffects::ggpredict(model = fit_polr_2,
                    terms = c("female")) 
```

```{r}
ggeffects::ggpredict(model = fit_polr_2,
                    terms = c("female")) %>% 
  data.frame()
```



## Hoffmann's Example 4.8 (continues...approximated)


> `ggeffects::ggpredict()`
> Adjusted for:
> *    female = male  *The reference category*
> * polviewsN = 4.17  *The grand mean value*


```{r}
ggeffects::ggpredict(model = fit_polr_2,
                    terms = c("educate [10, 16]",    # 1st = x
                              "female")) %>%         # 2nd = group
  data.frame() %>% 
  dplyr::filter(group == "male")
```

> `ggeffects::ggeffect()`
> Adjusted for:
> *    female *computed a kind of "average" value, which represents the proportions of male/female*
> * polviewsN = 4.17  *The grand mean value*

```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate [10, 16]")) %>% 
  data.frame()
```

```{r}
ggeffects::ggemmeans(model = fit_polr_2,
                     terms = c("educate [10, 16]"),
                     condition = c(female = "female")) %>% 
  data.frame()
```




Predictions for specific values:  females with 10 or 16 years education

```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("female",                 # 1st var = `x`
                              "educate [10, 16]")) %>%  # 2nd var = `group`
  data.frame()
```




```{r}
ggeffects::ggemmeans(model = fit_polr_2,
                     terms = "female",
                     condition = c(educate = 12,
                                   polviewsN  = 4.5)) 
```



### Plot Predicted Probabilites

```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate [10, 16]",   # x-axis
                              "female")) %>%        # lines by group
  data.frame() %>%  
  ggplot(aes(x = x,
             y = predicted,
             color = group,
             shape = group)) +
  geom_point(size = 4) +
  geom_line(aes(linetype = group)) +
  facet_wrap(~ response.level)
```



```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate [10, 16]",   # x-axis
                              "female")) %>%        # lines by group
  data.frame() %>% 
  dplyr::filter(response.level == "strongly.agree") %>% 
  ggplot(aes(x = x,
             y = predicted,
             color = group)) +
  geom_point(size = 4) +
  geom_line(aes(linetype = group)) 
```


```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate [10, 16]",   # x-axis
                              "female")) %>%        # lines by group
  data.frame() %>% 
  dplyr::filter(response.level == "strongly.agree") %>% 
  ggplot(aes(x = x,
             y = predicted,
             color = group,
             shape = group)) +
  geom_errorbar(aes(ymin = conf.low,
                    ymax = conf.high),
                width = .5,
                position = position_dodge(width =.25)) +
  geom_point(size = 4,
             position = position_dodge(width =.25)) +
  geom_line(aes(linetype = group),
            position = position_dodge(width =.25)) +
  theme_bw() +
  labs(x = "Education, years",
       y = "Predicted Probability for Strongly Agree",
       color = NULL,
       shape = NULL,
       linetype = NULL)
```


```{r, fig.cap="Hoffmann's Figure 4.4"}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate [10, 16]",   # x-axis
                              "female")) %>%        # lines by group
  data.frame() %>% 
  dplyr::mutate(group = forcats::fct_rev(group)) %>% 
  dplyr::filter(response.level == "strongly.agree") %>% 
  ggplot(aes(x = x,
             y = predicted,
             shape = group)) +
  geom_errorbar(aes(ymin = conf.low,
                    ymax = conf.high),
                width = .25,
                position = position_dodge(.2)) +
  geom_point(size = 4,
                position = position_dodge(.2)) +
  geom_line(aes(linetype = group),
            size = 1,
            position = position_dodge(.2)) +
  theme_bw() + 
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.key.width = unit(2, "cm"),
        legend.background = element_rect(color = "black")) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  labs(x = "Years of Formal Education",
       y = "Predicted Probabilit for\nResponding 'Strongly Agree'",
       color = NULL,
       shape = NULL,
       linetype = NULL,
       title = "Adjusted Predictions: Strongly Agree Spanking is Appropriate")
```



```{r}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("female")) %>%        # lines by group
  data.frame() %>% 
  dplyr::filter(response.level %in% c("strongly.agree",
                                      "strongly.disagree")) %>% 
  dplyr::mutate(resonse.level = factor(response.level)) %>% 
  ggplot(aes(x = x,
             y = predicted,
             fill = resonse.level)) +
  geom_col(position = position_dodge()) 

```


```{r, fig.cap = "Hoffmann's Figure 4.5"}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("female")) %>%        # lines by group
  data.frame() %>% 
  dplyr::filter(response.level %in% c("strongly.agree",
                                      "strongly.disagree")) %>% 
  dplyr::mutate(resonse.level = factor(response.level)) %>% 
  ggplot(aes(x = forcats::fct_rev(x),
             y = predicted,
             fill = resonse.level)) +
  geom_col(position = position_dodge()) +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("gray30", "gray70")) +
  labs(x = NULL,
       y = "Predicted Probability of Response",
       fill = NULL,
       title = "Attitues towareds Spanking, by Sex")

```


```{r, include=FALSE}
ggeffects::ggeffect(model = fit_polr_2,
                    terms = c("educate",                            # 1st = x
                              "female",                             # 2nd = group
                              "polviewsN [2.8, 4.2, 5.5]")) %>%  # 3rd = facet
  data.frame() %>% 
  dplyr::mutate(response.level = response.level %>% 
                  factor(levels = c("strongly.agree",
                                    "agree",
                                    "disagree",
                                    "strongly.disagree"))) %>% 
  dplyr::mutate(facet = facet %>% 
                  factor() %>% 
                  forcats::fct_recode("Low Polytical View (M - SD)"  = "2.8",
                                      "Mean Polytical View (M)"      = "4.2",
                                      "High Polytical View (M + SD)" = "5.5")) %>% 
  ggplot(aes(x = x,
             y = predicted,
             group = response.level,
             fill  = response.level)) +
  geom_ribbon(aes(ymin = conf.low,
                    ymax = conf.high),
                alpha = .25) + 
  geom_line(aes(linetype = response.level,
                color = response.level),
            size = 1) +
  theme_bw() +
  labs(x = NULL,
       y = "Predicted Probability",
       color    = "Spanking:",
       fill    = "Spanking:",
       linetype = "Spanking:") +
  theme(legend.position = "bottom",
        legend.key.width = unit(1.5, "cm")) +
  scale_linetype_manual(values = c("solid", "longdash", "dotdash", "dotted")) +
  scale_shape_manual(values = c(0, 1, 2, 8)) +
  facet_grid(group ~ facet)
```



### Model Fit and Variance Explained

```{r}
fit_polr_1redeo <- MASS::polr(spanking ~ female,
                         data = data_gss_model %>% 
                           dplyr::filter(complete.cases(educate, polviewsN)))
```



```{r}
anova(fit_polr_2, fit_polr_1redeo)
```

```{r}
performance::compare_performance(fit_polr_2, fit_polr_1redeo, rank = TRUE)
```





### Assumptions

#### Proportional Odds: Brant Test


The `poTest` function implements tests proposed by Brant (1990) for proportional odds for logistic models fit by the `polr()` function in the MASS package.

```{r}
# Hoffmann's Example 4.8
car::poTest(fit_polr_2)
```

A significant test statistics provides evidence that the parallel regression assumption has been violated!


